---
title: "CAZyme fold architecture is conserved between disparate environments despite extreme sequence divergence"
subtitle: "Oliyad Jeilu, Addis Simachew, Erica M. Hartmann, Erik Alexandersson, Eva Johansson"
author: "Oliyad Jeilu, Addis Simachew, Erica M. Hartmann, Erik Alexandersson, Eva Johansson | Published in mSystems (2026) | DOI: 10.1128/msystems.00485-26"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    theme: cosmo
    code_folding: hide
    fig_width: 10
    fig_height: 7
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", dpi = 300,
  dev = c("png", "pdf"),
  cache = FALSE
)

# ── Install / load packages ──────────────────────────────────────────────────
required_pkgs <- c(
  "tidyverse", "vegan", "ComplexHeatmap", "RColorBrewer", "ggtree", "treeio",
  "ape", "ggpubr", "rstatix", "dunn.test", "patchwork", "viridis",
  "scales", "kableExtra", "ggrepel", "reshape2", "ggsci", "circlize", "png"
)

for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg %in% c("ggtree", "treeio", "ComplexHeatmap")) {
      if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
      BiocManager::install(pkg, ask = FALSE, update = FALSE)
    } else {
      install.packages(pkg, repos = "https://cloud.r-project.org")
    }
  }
}

library(tidyverse); library(vegan); library(ComplexHeatmap); library(RColorBrewer)
library(ggtree); library(treeio); library(ape); library(ggpubr)
library(rstatix); library(dunn.test); library(patchwork); library(viridis)
library(scales); library(kableExtra); library(ggrepel); library(reshape2)
library(ggsci); library(circlize)

# ── Font fallback ────────────────────────────────────────────────────────────
# "sans" maps to Arial (Windows), Helvetica (Mac), DejaVu Sans (Linux)
# Avoids "invalid font type" error when Arial isn't registered in R
pub_font <- "sans"

# ── Global variables ─────────────────────────────────────────────────────────
data_dir <- "."

# ══════════════════════════════════════════════════════════════════════════════
# NATURE MICROBIOLOGY – FIGURE SPECIFICATIONS
# ══════════════════════════════════════════════════════════════════════════════
fig_dir <- "figures"
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)

nm_single <- 88 / 25.4    # 3.46 in
nm_1.5    <- 120 / 25.4   # 4.72 in
nm_full   <- 180 / 25.4   # 7.09 in
nm_max_h  <- 230 / 25.4   # 9.06 in
nm_wide   <- 320 / 25.4   # 12.60 in — wide multi-panel composites
nm_tall   <- 300 / 25.4   # 11.81 in — tall multi-panel composites

# Helper: save ggplot as PDF + PNG at 300 DPI
save_fig <- function(p, name, w, h, dpi = 300) {
  ggsave(file.path(fig_dir, paste0(name, ".pdf")), p, width = w, height = h,
         units = "in", device = cairo_pdf)
  ggsave(file.path(fig_dir, paste0(name, ".png")), p, width = w, height = h,
         units = "in", dpi = dpi)
  ggsave(file.path(fig_dir, paste0(name, ".tif")), p, width = w, height = h,
         units = "in", dpi = dpi, device = "tiff", compression = "lzw")
}

# Helper: save ComplexHeatmap as PDF + PNG
save_complexheatmap <- function(ht, name, w, h, dpi = 300) {
  pdf(file.path(fig_dir, paste0(name, ".pdf")), width = w, height = h)
  draw(ht); dev.off()
  png(file.path(fig_dir, paste0(name, ".png")), width = w, height = h,
      units = "in", res = dpi)
  draw(ht); dev.off()
  tiff(file.path(fig_dir, paste0(name, ".tif")), width = w, height = h,
       units = "in", res = dpi, compression = "lzw")
  draw(ht); dev.off()
}

# NPG palette
npg_cols <- pal_npg("nrc")(10)
env_levels  <- c("Soda Lake", "Rumen")
env_colors  <- c("Soda Lake" = npg_cols[1], "Rumen" = npg_cols[4])
source_colors <- c(
  "Abijata" = "#E64B35", "Chitu" = "#F39B7F", "Shala" = "#DC0000",
  "Goat" = "#4DBBD5", "Cattle" = "#00A087", "Sheep" = "#3C5488"
)
phylum_pal <- c(pal_d3("category20")(20), "grey60")

# ── Publication-quality ggplot2 theme (Nature Microbiology) ──────────────────
theme_pub <- function(base_size = 8, base_family = pub_font) {
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      text = element_text(colour = "black", family = base_family),
      plot.title = element_text(size = base_size + 1, face = "bold",
                                hjust = 0, margin = margin(b = 4)),
      plot.subtitle = element_text(size = base_size - 1, colour = "grey30",
                                   hjust = 0, margin = margin(b = 3)),
      axis.title = element_text(size = base_size, face = "bold"),
      axis.text = element_text(size = base_size - 1, colour = "black"),
      legend.title = element_text(size = base_size - 0.5, face = "bold"),
      legend.text = element_text(size = base_size - 1),
      strip.text = element_text(size = base_size, face = "bold"),
      axis.line = element_line(colour = "black", linewidth = 0.4),
      axis.ticks = element_line(colour = "black", linewidth = 0.3),
      axis.ticks.length = unit(0.12, "cm"),
      panel.grid = element_blank(), panel.border = element_blank(),
      panel.background = element_blank(),
      strip.background = element_rect(fill = "grey95", colour = NA),
      legend.background = element_blank(), legend.key = element_blank(),
      legend.key.size = unit(0.35, "cm"), legend.position = "right",
      plot.margin = margin(6, 6, 6, 6)
    )
}

# ── Helper: assign environment & source from sample IDs ──────────────────────
assign_env <- function(x) {
  env_chr <- case_when(
    str_detect(x, "^(ABS|AS[^a-z]|ASP)") ~ "Soda Lake",
    str_detect(x, "^(CHS|CW|CSP|CLP)")   ~ "Soda Lake",
    str_detect(x, "^(SHL|SWL|SLP)")       ~ "Soda Lake",
    str_detect(x, "^G")                   ~ "Rumen",
    str_detect(x, "^C")                   ~ "Rumen",
    str_detect(x, "^S")                   ~ "Rumen",
    TRUE                                  ~ NA_character_
  )
  factor(env_chr, levels = env_levels)
}

assign_source <- function(x) {
  case_when(
    str_detect(x, "^(ABS|AS[^a-z]|ASP)") ~ "Abijata",
    str_detect(x, "^(CHS|CW|CSP|CLP)")   ~ "Chitu",
    str_detect(x, "^(SHL|SWL|SLP)")       ~ "Shala",
    str_detect(x, "^G")                   ~ "Goat",
    str_detect(x, "^C")                   ~ "Cattle",
    str_detect(x, "^S")                   ~ "Sheep",
    TRUE                                  ~ NA_character_
  )
}
```

---

# 1 — Data Loading & Preprocessing 
## 1.1 Load all datasets

```{r load-data}
red_df <- read_tsv(file.path(data_dir, "gtdbtk_red_env.tsv"), show_col_types = FALSE) %>%
  mutate(RED = as.numeric(RED)) %>% filter(!is.na(RED))
if ("Environment" %in% names(red_df))
  red_df <- red_df %>% mutate(Environment = str_replace(Environment, "_", " "))

gtdbtk_full <- read_tsv(file.path(data_dir, "gtdbtk_bac120.tsv"), show_col_types = FALSE)

cazyme_2t <- read_tsv(file.path(data_dir, "all_cazymes_2tools_env.tsv"), show_col_types = FALSE) %>%
  mutate(Environment = str_replace(Environment, "_", " "))

diamond_df <- read_tsv(file.path(data_dir, "all_diamond_merged_env.tsv"), show_col_types = FALSE) %>%
  mutate(Environment = str_replace(Environment, "_", " "))

checkm_drep <- read_csv(file.path(data_dir, "checkm_for_drep_MQ.csv"), show_col_types = FALSE)

metaphlan <- read_tsv(file.path(data_dir, "metaphlan_merged.tsv"),
                      comment = "#", show_col_types = FALSE)

tree_backbone <- read.tree(file.path(data_dir, "gtdbtk_backbone.tree"))

prokka_ann <- read_tsv(file.path(data_dir, "prokka_drep_annotations.tsv"), show_col_types = FALSE) %>%
  mutate(Environment = assign_env(MAG), Sample_Source = assign_source(MAG)) %>%
  filter(!is.na(Environment))
prokka_sum <- read_tsv(file.path(data_dir, "prokka_drep_summary.tsv"), show_col_types = FALSE) %>%
  mutate(Environment = assign_env(MAG), Sample_Source = assign_source(MAG)) %>%
  filter(!is.na(Environment))
```

## 1.2 Parse taxonomy & build helper columns
#parssing. the taxa 

```{r parse-taxonomy}
parse_gtdbtk_taxonomy <- function(classification) {
  ranks <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  parts <- str_split(classification, ";", simplify = FALSE)
  map_dfr(parts, function(p) {
    vals <- str_remove(p, "^[dpcofgs]__")
    out  <- setNames(vals[1:7], ranks)
    as_tibble_row(out)
  })
}

red_tax <- red_df %>% bind_cols(parse_gtdbtk_taxonomy(red_df$Classification))

extract_cazyme_class <- function(hmmer) {
  fam <- str_extract(hmmer, "^[A-Z]+\\d*")
  case_when(
    str_starts(fam, "GH") ~ "GH", str_starts(fam, "GT") ~ "GT",
    str_starts(fam, "PL") ~ "PL", str_starts(fam, "CE") ~ "CE",
    str_starts(fam, "AA") ~ "AA", str_starts(fam, "CBM") ~ "CBM",
    TRUE ~ "Other"
  )
}

cazyme_2t <- cazyme_2t %>%
  mutate(
    CAZyme_Family = str_extract(HMMER, "^[A-Z]+\\d+(_\\d+)?"),
    CAZyme_Class  = extract_cazyme_class(HMMER)
  )

# Parse novelty ranks (needed for Figures 2, 3, 6)
novelty_rank <- gtdbtk_full %>%
  mutate(Environment = assign_env(user_genome)) %>%
  filter(!is.na(Environment)) %>%
  separate(classification, into = c("d","p","c","o","f","g","s"), sep = ";",
           remove = FALSE) %>%
  mutate(
    across(c(d,p,c,o,f,g,s), ~ str_remove(., "^[dpcofgs]__")),
    novel_species = (s == ""), novel_genus = (g == ""), novel_family = (f == "")
  )
```

---

# 2 — Community Composition — Figure 1 
## 2.1 Parse MetaPhlAn & assign metadata
#assinging the metdata from tehmetaphlan to metdata 

```{r mpa-parse}
colnames(metaphlan)[1] <- "clade_name"
colnames(metaphlan)[-1] <- str_remove(colnames(metaphlan)[-1], "_shotgun$")
metaphlan <- metaphlan %>% mutate(depth = str_count(clade_name, "\\|") + 1)

mpa_samples <- tibble(Sample = colnames(metaphlan)[-c(1, ncol(metaphlan))]) %>%
  mutate(Environment = assign_env(Sample), Source = assign_source(Sample))
```

## 2.2 Panel A — Phylum-level Community Composition
#here we are interested to identiphy the top phlum we used realtive abundance

```{r fig1-panelA, fig.height=7, fig.width=12}
#check why deoth 2 done here if imoiratnt 
mpa_phylum <- metaphlan %>%
  filter(depth == 2) %>%
  mutate(Phylum = str_extract(clade_name, "p__[A-Za-z_]+") %>% str_remove("p__"))
#uncalssified ? unassined ,,, 
mpa_phy_long <- mpa_phylum %>%
  select(-clade_name, -depth) %>%
  pivot_longer(-Phylum, names_to = "Sample", values_to = "Abundance") %>%
  left_join(mpa_samples, by = "Sample") %>%
  filter(!is.na(Phylum),
         !str_detect(Phylum, regex("unclassified|unmapped", ignore_case = TRUE)),
         Phylum != "")

top_phy <- mpa_phy_long %>%
  group_by(Phylum) %>% summarise(total = sum(Abundance), .groups = "drop") %>%
  slice_max(total, n = 12) %>% pull(Phylum)

mpa_phy_plot <- mpa_phy_long %>%
  mutate(Phylum = ifelse(Phylum %in% top_phy, Phylum, "Other")) %>%
  group_by(Sample, Phylum, Environment, Source) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

sample_order <- mpa_samples %>%
  mutate(env_rank = ifelse(Environment == "Soda Lake", 1, 2)) %>%
  arrange(env_rank, Source, Sample) %>% pull(Sample)
mpa_phy_plot$Sample <- factor(mpa_phy_plot$Sample, levels = sample_order)

p_phy_bar <- ggplot(mpa_phy_plot, aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", position = "fill", width = 0.85) +
  scale_fill_manual(values = phylum_pal) +
  scale_y_continuous(labels = percent_format()) +
  facet_grid(~ Environment, scales = "free_x", space = "free_x") +
  labs(y = "Relative Abundance (%)", x = NULL) +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 6),
        strip.text = element_text(size = 8))
p_phy_bar
```

## 2.3 Panel B — Top 30 Genera Heatmap (ComplexHeatmap)
#okey now kets go to teh txaa deep we have to look for genus and top 30are the one we need to look for 
```{r fig1-panelB, fig.height=10, fig.width=14}
# Species-level matrix (needed later for alpha diversity & PCoA)
# Build carefully: species (rows) × samples (cols), then transpose
mpa_sp_dat <- metaphlan %>% filter(depth == 7)
sp_clade_names <- mpa_sp_dat$clade_name
sp_mat <- mpa_sp_dat %>% select(-clade_name, -depth) %>% as.data.frame()
rownames(sp_mat) <- sp_clade_names
sp_mat <- as.matrix(sp_mat)
storage.mode(sp_mat) <- "double"         # force numeric
sp_mat_t <- t(sp_mat)                    # samples (rows) × species (cols)
cat("sp_mat_t dimensions:", nrow(sp_mat_t), "samples ×", ncol(sp_mat_t), "species\n")

# Extract genus-level data
mpa_genus <- metaphlan %>%
  filter(depth == 6) %>%
  mutate(Genus = str_extract(clade_name, "g__[^|]+") %>% str_remove("g__"))
#why depth 6 is ookisng here , but phylum wa depth 2 ?
# ── CLEAN genus names: remove artefacts ──────────────────────────────────────
# Filter out: unmapped, unclassified, "Other", weird codes (sggg, ggB, single
# letters/numbers, names shorter than 3 chars, GTDB placeholder codes)
mpa_genus_clean <- mpa_genus %>%
  filter(
    !is.na(Genus), Genus != "",
    !str_detect(Genus, regex("unclassified|unmapped|other|unknown", ignore_case = TRUE)),
    nchar(Genus) >= 3,
    !str_detect(Genus, "^[A-Z]{1,3}\\d"),       # GGB1234, SGB123
    !str_detect(Genus, "^[a-z]{2,4}$"),         # weird short codes
    !str_detect(Genus, "^\\d+$"),               # pure numbers
    !str_detect(Genus, "^[A-Z][A-Z]+\\d"),      # JAGJPN01 etc
    !str_detect(Genus, "^[A-Z]{2,}_")           # UBA1234, CAG_xyz
  )

genus_mat <- mpa_genus_clean %>%
  select(-clade_name, -depth) %>%
  column_to_rownames("Genus") %>% as.matrix()

# Keep genera in ≥3 samples with mean > 0.01%
keep_genus <- rownames(genus_mat)[rowSums(genus_mat > 0) >= 3 & rowMeans(genus_mat) > 0.01]
genus_filt <- genus_mat[keep_genus, ]

# Top 30
top30_idx <- sort(rowMeans(genus_filt), decreasing = TRUE)
if (length(top30_idx) > 30) top30_idx <- top30_idx[1:30]
genus_filt <- genus_filt[names(top30_idx), ]

# Log-transform
genus_log <- log2(genus_filt + 0.001)

# ── ComplexHeatmap with metadata annotations ─────────────────────────────────
# Column (sample) annotation
col_meta <- mpa_samples %>% filter(Sample %in% colnames(genus_log))
col_meta <- col_meta[match(colnames(genus_log), col_meta$Sample), ]

ha_top <- HeatmapAnnotation(
  Environment = col_meta$Environment,
  Source = col_meta$Source,
  col = list(
    Environment = env_colors,
    Source = source_colors
  ),
  annotation_name_gp = gpar(fontsize = 7, fontfamily = pub_font),
  simple_anno_size = unit(3, "mm"),
  show_legend = TRUE
)

# Colour scale
col_fun <- colorRamp2(
  c(min(genus_log), 0, max(genus_log)),
  c("#2166AC", "white", "#B2182B")
)

ht_genus <- Heatmap(
  genus_log,
  name = "log2\nabundance",
  col = col_fun,
  top_annotation = ha_top,
  row_names_gp = gpar(fontsize = 7, fontface = "italic", fontfamily = pub_font),
  column_names_gp = gpar(fontsize = 6, fontfamily = pub_font),
  column_names_rot = 50,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  row_dend_width = unit(15, "mm"),
  column_dend_height = unit(15, "mm"),
  column_split = col_meta$Environment,
  column_title_gp = gpar(fontsize = 8, fontface = "bold", fontfamily = pub_font),
  border = FALSE,
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 7, fontface = "bold", fontfamily = pub_font),
    labels_gp = gpar(fontsize = 6, fontfamily = pub_font),
    legend_height = unit(3, "cm")
  )
)
draw(ht_genus, column_title = "Top 30 Genera — MetaPhlAn4 (log2-transformed)",
     column_title_gp = gpar(fontsize = 9, fontface = "bold", fontfamily = pub_font))
# Save standalone copy
save_complexheatmap(ht_genus, "Fig1B_genus_heatmap", nm_wide, nm_tall)

# Capture as grob for patchwork integration
p_genus_ht <- wrap_elements(full = grid::grid.grabExpr(
  draw(ht_genus, column_title = "Top 30 Genera (log2-transformed)",
       column_title_gp = gpar(fontsize = 8, fontface = "bold", fontfamily = pub_font)),
  width = 14, height = 12
))
```

## 2.4 Panel C — Unclassified Fraction
# oke the next quesion we wanna adress is the uncassified version but we are lookng for metaphlan 

```{r fig1-panelC, fig.height=5, fig.width=6}
unclass_row <- metaphlan %>% filter(clade_name == "UNCLASSIFIED")
unclass_df <- unclass_row %>%
  select(-clade_name, -depth) %>%
  pivot_longer(everything(), names_to = "Sample", values_to = "Pct_Unclassified") %>%
  left_join(mpa_samples, by = "Sample")

wil_uc <- wilcox.test(Pct_Unclassified ~ Environment, data = unclass_df)

p_unclass <- ggplot(unclass_df, aes(x = Environment, y = Pct_Unclassified, fill = Environment)) +
  geom_boxplot(alpha = 0.75, width = 0.5, outlier.shape = NA) +
  geom_jitter(aes(colour = Source), width = 0.12, size = 2.5, alpha = 0.85) +
  scale_fill_manual(values = env_colors) +
  scale_colour_manual(values = source_colors) +
  labs(subtitle = sprintf("Wilcoxon p = %.3e", wil_uc$p.value),
       y = "Unclassified (%)", x = NULL, colour = "Source") +
  theme_pub() + guides(fill = "none")
p_unclass
```

## 2.5 Panel D — Alpha Diversity

```{r fig1-panelD, fig.height=6, fig.width=11}
# Compute per-sample (row) metrics explicitly with MARGIN = 1
obs_vec <- vegan::specnumber(sp_mat_t, MARGIN = 1)
sha_vec <- vegan::diversity(sp_mat_t, index = "shannon", MARGIN = 1)
mpa_alpha <- tibble(
  Sample = rownames(sp_mat_t),
  Observed = as.numeric(obs_vec),
  Shannon = as.numeric(sha_vec)
) %>% left_join(mpa_samples, by = "Sample")

kw_obs <- kruskal.test(Observed ~ Environment, data = mpa_alpha)
kw_sha <- kruskal.test(Shannon ~ Environment, data = mpa_alpha)

p_obs <- ggplot(mpa_alpha, aes(x = Environment, y = Observed, fill = Environment)) +
  geom_boxplot(alpha = 0.7, width = 0.5, outlier.shape = NA) +
  geom_jitter(aes(colour = Source), width = 0.12, size = 2, alpha = 0.8) +
  scale_fill_manual(values = env_colors) +
  scale_colour_manual(values = source_colors) +
  labs(title = "Observed Species", subtitle = sprintf("KW p = %.2e", kw_obs$p.value),
       y = "Number of species", x = NULL) +
  theme_pub() + guides(fill = "none")

p_sha <- ggplot(mpa_alpha, aes(x = Environment, y = Shannon, fill = Environment)) +
  geom_boxplot(alpha = 0.7, width = 0.5, outlier.shape = NA) +
  geom_jitter(aes(colour = Source), width = 0.12, size = 2, alpha = 0.8) +
  scale_fill_manual(values = env_colors) +
  scale_colour_manual(values = source_colors) +
  labs(title = "Shannon Diversity", subtitle = sprintf("KW p = %.2e", kw_sha$p.value),
       y = "Shannon Index (H')", x = NULL) +
  theme_pub() + guides(fill = "none")

p_alpha_combined <- p_obs + p_sha + plot_layout(guides = "collect") &
  theme(legend.position = "right")
p_alpha_combined
```

## 2.6 Panel E — PCoA (Bray-Curtis β-diversity)
# now lets cehck if the structural composition difeerent between these two ecosytems 

```{r fig1-panelE, fig.height=7, fig.width=9}
bc_mpa <- vegdist(sp_mat_t, method = "bray")
pcoa_res <- cmdscale(bc_mpa, k = 2, eig = TRUE)
eig_pct <- round(100 * pcoa_res$eig / sum(pcoa_res$eig[pcoa_res$eig > 0]), 1)
#braycurtsis

pcoa_df <- tibble(
  Sample = rownames(pcoa_res$points),
  PCoA1 = pcoa_res$points[, 1], PCoA2 = pcoa_res$points[, 2]
) %>% left_join(mpa_samples, by = "Sample")

# Ensure sample order matches distance matrix
perm_meta <- mpa_samples %>% filter(Sample %in% rownames(sp_mat_t))
perm_meta <- perm_meta[match(rownames(sp_mat_t), perm_meta$Sample), ]
perm_mpa <- adonis2(bc_mpa ~ Environment, data = perm_meta, permutations = 999)

p_pcoa <- ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, colour = Source, shape = Environment)) +
  geom_point(size = 2.5, alpha = 0.85) +
  stat_ellipse(aes(group = Environment, colour = NULL), linetype = "dashed",
               linewidth = 0.5, colour = "grey40") +
  scale_colour_manual(values = source_colors) +
  scale_shape_manual(values = c("Soda Lake" = 16, "Rumen" = 17)) +
  labs(subtitle = sprintf("PERMANOVA R² = %.3f, p = %.3f", perm_mpa$R2[1], perm_mpa$`Pr(>F)`[1]),
       x = sprintf("PCoA1 (%.1f%%)", eig_pct[1]),
       y = sprintf("PCoA2 (%.1f%%)", eig_pct[2]),
       colour = "Source", shape = "Environment") +
  theme_pub()
p_pcoa
```

## 2.7 ★ Figure 1: Community Composition (A–E)
#okey now lets merge and make ccompsute 
```{r fig1-composite, fig.height=11.81, fig.width=12.60}
# Panel order: A = Phylum bar, B = Genus heatmap, C = Unclassified, D = Alpha, E = PCoA
# Wrap alpha_combined so patchwork sees it as ONE panel (not two sub-panels)
p_alpha_wrap <- wrap_elements(full = p_alpha_combined)

p_fig1 <- (p_phy_bar + labs(title = NULL)) /
           p_genus_ht /
           (p_unclass + labs(title = NULL, subtitle = NULL) |
            p_alpha_wrap) /
           (p_pcoa + labs(title = NULL, subtitle = NULL)) +
  plot_layout(heights = c(1, 2.2, 1, 1)) +
  plot_annotation(
    tag_levels = "A",
    title = ""
  ) &
  theme_pub() &
  theme(plot.tag = element_text(size = 10, face = "bold"),
        plot.margin = margin(4, 4, 4, 4))
p_fig1
save_fig(p_fig1, "Fig1_community_composition", nm_wide, nm_tall)
```

---

# 3 — Phylogenetic Novelty (RED, ANI) — Figure 2 
#here i used mags .. the assembly it was used megahit , metabat2 binning 
## 3.1 Panel A — RED: Soda Lake vs Rumen

```{r fig2-panelA, fig.height=5.5, fig.width=6.5}
kw_env <- kruskal.test(RED ~ Environment, data = red_df)

p_red_env <- ggplot(red_df, aes(x = Environment, y = RED, fill = Environment)) +
  geom_violin(alpha = 0.3, width = 0.8, colour = NA) +
  geom_boxplot(alpha = 0.8, width = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.2, size = 0.5) +
  scale_fill_manual(values = env_colors) +
  stat_compare_means(method = "wilcox.test", label = "p.signif",
                     label.x = 1.5, label.y = max(red_df$RED, na.rm = TRUE) * 1.02,
                     size = 4, bracket.size = 0.4) +
  labs(subtitle = sprintf("Kruskal-Wallis p = %.2e", kw_env$p.value),
       y = "Relative Evolutionary Divergence (RED)", x = NULL) +
  theme_pub() + theme(legend.position = "none")
p_red_env
```

## 3.2 Panel B — ANI to Closest Reference
# AnI for novelity check .. ANI vs RD ?/ 

```{r fig2-panelB, fig.height=6, fig.width=10}
gtdbtk_ani <- gtdbtk_full %>%
  mutate(
    Environment = assign_env(user_genome), Source = assign_source(user_genome),
    ANI = as.numeric(closest_genome_ani), AF = as.numeric(closest_genome_af)
  ) %>% filter(!is.na(Environment), !is.na(ANI))

kw_ani <- kruskal.test(ANI ~ Environment, data = gtdbtk_ani)

p_ani <- ggplot(gtdbtk_ani, aes(x = Environment, y = ANI, fill = Environment)) +
  geom_violin(alpha = 0.3, colour = NA) +
  geom_boxplot(alpha = 0.75, width = 0.3, outlier.shape = NA) +
  geom_jitter(aes(colour = Source), width = 0.08, size = 1.5, alpha = 0.6) +
  geom_hline(yintercept = 95, linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  annotate("text", x = 0.6, y = 95.3, label = "Species threshold (95%)",
           size = 2.5, colour = "grey40", hjust = 0) +
  scale_fill_manual(values = env_colors) +
  scale_colour_manual(values = source_colors) +
  labs(subtitle = sprintf("KW p = %.2e", kw_ani$p.value),
       y = "ANI (%)", x = NULL, colour = "Source") +
  theme_pub() + guides(fill = "none")
p_ani
```

## 3.3 Panel C — Taxonomic Novelty by Rank
#novelity is based on assembled MAGS but no hit 

```{r fig2-panelC, fig.height=6, fig.width=10}
novelty_long <- novelty_rank %>%
  select(user_genome, Environment, novel_species, novel_genus, novel_family) %>%
  pivot_longer(starts_with("novel_"), names_to = "Rank", values_to = "is_novel") %>%
  mutate(
    Rank = str_remove(Rank, "novel_") %>% str_to_title(),
    Rank = factor(Rank, levels = c("Family", "Genus", "Species")),
    Status = ifelse(is_novel, "Novel", "Classified")
  )

p_novelty_rank <- ggplot(novelty_long, aes(x = Rank, fill = Status)) +
  geom_bar(position = "fill", width = 0.65) +
  scale_fill_manual(values = c("Classified" = npg_cols[4], "Novel" = npg_cols[1])) +
  scale_y_continuous(labels = percent_format()) +
  facet_grid(~ Environment) +
  labs(x = NULL, y = "Proportion of MAGs") +
  theme_pub()
p_novelty_rank
```

## 3.4 Panel D — CheckM Quality Scatter

```{r fig2-panelD, fig.height=6, fig.width=8}
checkm_plot <- checkm_drep %>%
  mutate(
    Environment = assign_env(genome), Source = assign_source(genome),
    Quality = case_when(
      completeness >= 90 & contamination <= 5 ~ "High",
      completeness >= 50 & contamination <= 10 ~ "Medium",
      TRUE ~ "Low"
    )
  )

p_checkm <- ggplot(checkm_plot, aes(x = completeness, y = contamination, colour = Environment)) +
  geom_point(alpha = 0.7, size = 1.5) +
  geom_hline(yintercept = c(5, 10), linetype = "dashed", colour = "grey60", linewidth = 0.3) +
  geom_vline(xintercept = c(50, 90), linetype = "dashed", colour = "grey60", linewidth = 0.3) +
  scale_colour_manual(values = env_colors) +
  annotate("rect", xmin = 90, xmax = 100, ymin = 0, ymax = 5, alpha = 0.08, fill = "green") +
  annotate("text", x = 95, y = 4, label = "HQ", fontface = "bold", size = 2.5) +
  labs(x = "Completeness (%)", y = "Contamination (%)") +
  theme_pub()
p_checkm
```

## 3.5 Panel E — MAG Phylum Composition

```{r fig2-panelE, fig.height=7, fig.width=11}
phylum_counts <- red_tax %>%
  filter(Phylum != "") %>%
  count(Sample_Source, Environment, Phylum, name = "MAG_count")

top12_phyla <- phylum_counts %>%
  group_by(Phylum) %>% summarise(total = sum(MAG_count), .groups = "drop") %>%
  slice_max(total, n = 12) %>% pull(Phylum)

p_mag_phy <- phylum_counts %>%
  mutate(Phylum = ifelse(Phylum %in% top12_phyla, Phylum, "Other")) %>%
  group_by(Sample_Source, Environment, Phylum) %>%
  summarise(MAG_count = sum(MAG_count), .groups = "drop") %>%
  ggplot(aes(x = Sample_Source, y = MAG_count, fill = Phylum)) +
  geom_bar(stat = "identity", position = "fill", width = 0.75) +
  scale_fill_manual(values = phylum_pal) +
  scale_y_continuous(labels = percent_format()) +
  facet_grid(~ Environment, scales = "free_x", space = "free_x") +
  labs(x = NULL, y = "Proportion") +
  theme_pub() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_mag_phy
```

## 3.6 Panel F — MAG Genus Composition (GTDB-Tk)

```{r fig2-panelF, fig.height=7, fig.width=12}
# Build genus-level composition from GTDB-Tk
mag_genus_comp <- novelty_rank %>%
  mutate(
    Source = assign_source(user_genome),
    Genus = case_when(
      g != "" ~ g,
      f != "" ~ paste0("Novel (", f, ")"),
      TRUE    ~ "Unclassified"
    )
  ) %>%
  filter(!is.na(Environment)) %>%
  # Clean genus names
  filter(
    !str_detect(Genus, regex("unclassified|unmapped", ignore_case = TRUE)),
    nchar(Genus) >= 3,
    !str_detect(Genus, "^[A-Z]{2,}_"),
    !str_detect(Genus, "^[A-Z]{2,}\\d")
  )

top15_genus <- mag_genus_comp %>%
  count(Genus, sort = TRUE) %>% slice_head(n = 15) %>% pull(Genus)

genus_pal_fig3 <- c(setNames(pal_d3("category20")(15), sort(top15_genus)), "Other" = "grey70")

p_mag_genus <- mag_genus_comp %>%
  mutate(Genus = ifelse(Genus %in% top15_genus, Genus, "Other")) %>%
  count(Source, Environment, Genus) %>%
  ggplot(aes(x = Source, y = n, fill = Genus)) +
  geom_bar(stat = "identity", position = "fill", width = 0.75) +
  scale_fill_manual(values = genus_pal_fig3) +
  scale_y_continuous(labels = percent_format()) +
  facet_grid(~ Environment, scales = "free_x", space = "free_x") +
  labs(x = NULL, y = "Proportion") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.text = element_text(face = "italic", size = 6),
        legend.key.size = unit(0.25, "cm")) +
  guides(fill = guide_legend(ncol = 2))
p_mag_genus
```

## 3.7 MAG Species Composition (GTDB-Tk)

```{r figS2-species, fig.height=8, fig.width=12}
# Build species-level composition from GTDB-Tk
mag_species <- novelty_rank %>%
  mutate(
    Source = assign_source(user_genome),
    Species = case_when(
      s != "" ~ s,
      g != "" ~ paste0(g, " sp."),
      f != "" ~ paste0("Novel (", f, ")"),
      TRUE    ~ "Unclassified"
    )
  ) %>%
  filter(!is.na(Environment))

# Clean species names: remove artefacts and GTDB placeholder codes
mag_species_clean <- mag_species %>%
  filter(
    !str_detect(Species, regex("unclassified|unmapped|unknown", ignore_case = TRUE)),
    !str_detect(Species, "^[A-Z]{2,}_"),
    !str_detect(Species, "^[A-Z]{2,}\\d"),
    nchar(Species) >= 3
  )

# Top 15 species WITHIN each environment (not overall)
top_per_env <- mag_species_clean %>%
  count(Environment, Species, sort = TRUE) %>%
  group_by(Environment) %>%
  slice_head(n = 15) %>%
  ungroup()

top_sp_union <- unique(top_per_env$Species)

sp_plot <- mag_species_clean %>%
  mutate(Species = ifelse(Species %in% top_sp_union, Species, "Other")) %>%
  count(Source, Environment, Species)

sp_pal <- c(setNames(
  colorRampPalette(pal_d3("category20")(20))(length(top_sp_union)),
  sort(top_sp_union)), "Other" = "grey70")

p_mag_species <- ggplot(sp_plot, aes(x = Source, y = n, fill = Species)) +
  geom_bar(stat = "identity", position = "fill", width = 0.75) +
  scale_fill_manual(values = sp_pal) +
  scale_y_continuous(labels = percent_format()) +
  facet_grid(~ Environment, scales = "free_x", space = "free_x") +
  labs(x = NULL, y = "Proportion of MAGs") +
  theme_pub() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.text = element_text(face = "italic", size = 5.5),
    legend.key.size = unit(0.25, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2))
p_mag_species
save_fig(p_mag_species, "FigS2_species_composition", nm_wide, nm_tall)
```

## 3.8 Panel G — RED vs Contamination + CAZyme vs Completeness (QC)

```{r fig2-panelG, fig.height=6, fig.width=12}
# RED vs contamination
red_contam <- red_df %>%
  mutate(genome_fa = paste0(MAG_ID, ".fa")) %>%
  left_join(checkm_drep %>% rename(genome_fa = genome), by = "genome_fa") %>%
  filter(!is.na(contamination), !is.na(RED))

cor_test <- cor.test(red_contam$RED, red_contam$contamination, method = "spearman")

p_red_contam <- ggplot(red_contam, aes(x = contamination, y = RED, colour = Environment)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey30", linewidth = 0.5, linetype = "dashed") +
  scale_colour_manual(values = env_colors) +
  labs(subtitle = sprintf("Spearman ρ = %.3f, p = %.3e", cor_test$estimate, cor_test$p.value),
       x = "Contamination (%)", y = "RED") +
  theme_pub()

# Build CAZyme alpha diversity (needed here and for Figure 4)
mag_cazyme <- cazyme_2t %>%
  filter(!is.na(CAZyme_Family)) %>%
  count(Bin_ID, Environment, Sample_Source, CAZyme_Family) %>%
  pivot_wider(names_from = CAZyme_Family, values_from = n, values_fill = 0)

fam_df <- mag_cazyme %>% select(-Bin_ID, -Environment, -Sample_Source) %>% as.data.frame()
rownames(fam_df) <- mag_cazyme$Bin_ID
fam_matrix <- as.matrix(fam_df)
storage.mode(fam_matrix) <- "double"

alpha_div <- tibble(
  Bin_ID = mag_cazyme$Bin_ID, Environment = mag_cazyme$Environment,
  Sample_Source = mag_cazyme$Sample_Source,
  Richness = as.numeric(vegan::specnumber(fam_matrix, MARGIN = 1)),
  Shannon  = as.numeric(vegan::diversity(fam_matrix, index = "shannon", MARGIN = 1)),
  Total_Genes = rowSums(fam_matrix)
)

alpha_qual <- alpha_div %>%
  mutate(genome_fa = paste0(Bin_ID, ".fa")) %>%
  left_join(checkm_drep %>% rename(genome_fa = genome), by = "genome_fa") %>%
  filter(!is.na(completeness))

cor_caz_comp <- cor.test(alpha_qual$Richness, alpha_qual$completeness, method = "spearman")

p_caz_qual <- ggplot(alpha_qual, aes(x = completeness, y = Richness, colour = Environment)) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey30", linewidth = 0.5, linetype = "dashed") +
  scale_colour_manual(values = env_colors) +
  labs(subtitle = sprintf("Spearman ρ = %.3f, p = %.3e", cor_caz_comp$estimate, cor_caz_comp$p.value),
       x = "Completeness (%)", y = "CAZyme Family Richness") +
  theme_pub()

p_red_contam | p_caz_qual
```

## 3.9 ★ Figure 2: MAG Characterisation — Quality, Novelty & Taxonomy

```{r fig2-composite, fig.height=14, fig.width=12.60}
# Panel flow: A=CheckM, B=RED, C=RED vs Contam | D=ANI, E=Novelty rank | F=Phylum | G=Species
p_fig2 <- (p_checkm + labs(title = NULL) |
            p_red_env + labs(title = NULL, subtitle = NULL) |
            p_red_contam + labs(title = NULL, subtitle = NULL)) /
           (p_ani + labs(title = NULL, subtitle = NULL) |
            p_novelty_rank + labs(title = NULL)) /
           (p_mag_phy + labs(title = NULL)) /
           (p_mag_species + labs(title = NULL)) +
  plot_layout(heights = c(1, 1, 1, 1)) +
  plot_annotation(
    tag_levels = "A",
    title = ""
  ) &
  theme_pub() &
  theme(plot.tag = element_text(size = 10, face = "bold"),
        plot.margin = margin(4, 4, 4, 4))
p_fig2
save_fig(p_fig2, "Fig2_mag_characterisation", nm_wide, 14)
```

---

# 4 — Functional & CAZyme Capacity — Figures 3–5 
## 4.1 Panel A — CAZyme Class Distribution

```{r fig3-panelA, fig.height=6, fig.width=11}
class_counts <- cazyme_2t %>%
  filter(!is.na(CAZyme_Class), CAZyme_Class != "Other") %>%
  count(Environment, Sample_Source, CAZyme_Class)

p_caz_class <- ggplot(class_counts, aes(x = Sample_Source, y = n, fill = CAZyme_Class)) +
  geom_bar(stat = "identity", position = "fill", width = 0.75) +
  scale_fill_manual(values = pal_npg("nrc")(6)) +
  scale_y_continuous(labels = percent_format()) +
  facet_grid(~ Environment, scales = "free_x", space = "free_x") +
  labs(x = NULL, y = "Proportion", fill = "CAZyme Class") +
  theme_pub() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_caz_class
```

## 4.2 Panel B — Per-MAG Richness & Shannon

```{r fig3-panelB, fig.height=6, fig.width=13}
kw_rich <- kruskal.test(Richness ~ Environment, data = alpha_div)
kw_shan <- kruskal.test(Shannon ~ Environment, data = alpha_div)

p_r <- ggplot(alpha_div, aes(x = Environment, y = Richness, fill = Environment)) +
  geom_violin(alpha = 0.25, colour = NA) +
  geom_boxplot(alpha = 0.7, width = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.15, size = 0.5) +
  scale_fill_manual(values = env_colors) +
  labs(subtitle = sprintf("KW p = %.2e", kw_rich$p.value),
       y = "Families per MAG", x = NULL) +
  theme_pub() + theme(legend.position = "none")

p_s <- ggplot(alpha_div, aes(x = Environment, y = Shannon, fill = Environment)) +
  geom_violin(alpha = 0.25, colour = NA) +
  geom_boxplot(alpha = 0.7, width = 0.3, outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.15, size = 0.5) +
  scale_fill_manual(values = env_colors) +
  labs(subtitle = sprintf("KW p = %.2e", kw_shan$p.value),
       y = "Shannon (H')", x = NULL) +
  theme_pub() + theme(legend.position = "none")

p_caz_alpha <- p_r + p_s
p_caz_alpha
```

## 4.3 Panel C — CAZyme Sequence Divergence (Identity Density)

```{r fig3-panelC, fig.height=6, fig.width=10}
diamond_env <- diamond_df %>%
  mutate(Environment = str_replace(Environment, "_", " "),
         Pct_Identity = as.numeric(Pct_Identity)) %>%
  filter(!is.na(Pct_Identity), !is.na(Environment))

wil_id <- wilcox.test(Pct_Identity ~ Environment, data = diamond_env)

novel_frac <- diamond_env %>%
  group_by(Environment) %>%
  summarise(n = n(), below_40 = sum(Pct_Identity < 40),
            pct_novel = 100 * below_40 / n, median_id = median(Pct_Identity), .groups = "drop")

p_caz_id <- ggplot(diamond_env, aes(x = Pct_Identity, fill = Environment)) +
  geom_density(alpha = 0.5, colour = NA) +
  geom_vline(xintercept = 40, linetype = "dashed", colour = "grey40", linewidth = 0.3) +
  annotate("text", x = 38, y = Inf, label = "< 40% = putatively novel",
           hjust = 1, vjust = 1.5, size = 2.5, colour = "grey40") +
  scale_fill_manual(values = env_colors) +
  labs(subtitle = sprintf("Wilcoxon p = %.2e | SL median = %.1f%% | Rumen median = %.1f%%",
                           wil_id$p.value,
                           novel_frac$median_id[novel_frac$Environment == "Soda Lake"],
                           novel_frac$median_id[novel_frac$Environment == "Rumen"]),
       x = "Sequence Identity (%)", y = "Density") +
  theme_pub()
p_caz_id
```

## 4.4 Build Functional Pathways (for Figure 3 composite)

```{r fig3-pathway-computation, include=FALSE}
# ── Define pathway-level gene groupings ──────────────────────────────────────
pathway_genes <- tribble(
  ~pathway,                       ~pathway_group,       ~genes,
  # ── pH homeostasis ──
  "Mrp (Na+/H+ antiporter)",     "pH homeostasis",     "mrpA,mrpB,mrpC,mrpD",
  "NhaA (Na+/H+ antiporter)",    "pH homeostasis",     "nhaA,nhaC",
  # ── Oxidative stress ──
  "KatG (catalase-peroxidase)",   "Oxidative stress",   "katG,katA",
  "SOD (superoxide dismutase)",   "Oxidative stress",   "sodA,sodB",
  "AhpC (peroxide reductase)",    "Oxidative stress",   "ahpC",
  "CspA (cold shock)",            "Oxidative stress",   "cspA,cspB",
  # ── Osmotic stress ──
  "Ect (ectoine biosynthesis)",   "Osmotic stress",     "ectA,ectB,ectC",
  "Bet (glycine betaine)",        "Osmotic stress",     "betA,betB",
  "ProU (betaine transport)",     "Osmotic stress",     "proV,proW,proX",
  "Kdp (K+ ATPase)",              "Osmotic stress",     "kdpA,kdpB",
  # ── Sulfur cycling ──
  "Sox (thiosulfate oxidation)",  "Sulfur cycling",     "soxA,soxB,soxC,soxY",
  "Dsr (dissimilatory sulfate)",  "Sulfur cycling",     "dsrA,dsrB",
  "Apr (sulfite oxidation)",      "Sulfur cycling",     "aprA,aprB",
  # ── Carbon fixation ──
  "RuBisCO (carbon fixation)",    "Carbon fixation",    "cbbL,cbbS,rbcL",
  # ── Nitrogen cycling ──
  "Nif (N2 fixation)",            "Nitrogen cycling",   "nifH,nifD,nifK",
  "Nir (nitrite reduction)",      "Nitrogen cycling",   "nirK,nirS",
  "Nar (nitrate reduction)",      "Nitrogen cycling",   "narG,narH",
  "NosZ (N2O reduction)",         "Nitrogen cycling",   "nosZ",
  "Nor (NO reduction)",           "Nitrogen cycling",   "norB",
  "Urease (urea hydrolysis)",     "Nitrogen cycling",   "ureA,ureB,ureC",
  # ── Carbohydrate catabolism ──
  "Cellulosome (scaffoldin)",     "Carbohydrate",       "cipA,cipB,cipC",
  "SusCD (polysaccharide util.)", "Carbohydrate",       "susC,susD",
  "CelA (endoglucanase)",         "Carbohydrate",       "celA,celB",
  "BglB (beta-glucosidase)",      "Carbohydrate",       "bglA,bglB",
  "Xyl (xylose utilization)",     "Carbohydrate",       "xylA,xylB",
  "GlgP (glycogen phosphoryl.)",  "Carbohydrate",       "glgP",
  # ── Fermentation ──
  "AckA (acetate production)",    "Fermentation",       "ackA,pta",
  "PflB (formate acetyltransf.)", "Fermentation",       "pflB",
  "PccB (propionate pathway)",    "Fermentation",       "pccB",
  # ── Metal resistance ──
  "ArsC (arsenate reductase)",    "Metal resistance",   "arsC"
) %>% mutate(gene_list = str_split(genes, ","))

group_levels <- c("pH homeostasis", "Oxidative stress", "Osmotic stress",
                  "Sulfur cycling", "Carbon fixation", "Nitrogen cycling",
                  "Carbohydrate", "Fermentation", "Metal resistance")
pathway_order <- pathway_genes %>%
  arrange(factor(pathway_group, levels = group_levels), pathway) %>%
  pull(pathway)

# ── Compute prevalence per pathway x source ──────────────────────────────────
n_mags_per_source <- prokka_ann %>%
  distinct(MAG, Sample_Source, Environment) %>%
  count(Sample_Source, Environment, name = "n_total")

n_sl_mags <- n_distinct(prokka_ann$MAG[prokka_ann$Environment == "Soda Lake"])
n_ru_mags <- n_distinct(prokka_ann$MAG[prokka_ann$Environment == "Rumen"])

hit_raw <- pathway_genes %>%
  unnest(gene_list) %>% rename(gene = gene_list) %>%
  inner_join(prokka_ann %>% select(MAG, gene, product, Sample_Source, Environment),
             by = "gene", relationship = "many-to-many") %>%
  # ── Remove known false-positive gene-name collisions ──────────────────────

  # soxA/soxB/soxD: "sarcosine oxidase" shares abbreviation with Sox thiosulfate oxidation

  # proX: "Prolyl-tRNA editing protein" shares abbreviation with betaine transport
  filter(!(gene %in% c("soxA","soxB","soxD") & str_detect(product, regex("sarcosine", ignore_case = TRUE))),
         !(gene == "proX" & str_detect(product, regex("prolyl-tRNA", ignore_case = TRUE)))) %>%
  select(-product) %>%
  group_by(pathway, pathway_group, MAG, Sample_Source, Environment) %>%
  summarise(gene_count = n_distinct(gene), .groups = "drop")

hit_summary <- hit_raw %>%
  group_by(pathway, pathway_group, Sample_Source, Environment) %>%
  summarise(n_mags_with = n_distinct(MAG),
            mean_genes  = mean(gene_count), .groups = "drop") %>%
  left_join(n_mags_per_source, by = c("Sample_Source", "Environment")) %>%
  mutate(prevalence = n_mags_with / n_total * 100)

full_grid <- expand_grid(
  pathway_genes %>% select(pathway, pathway_group),
  n_mags_per_source %>% select(Sample_Source, Environment)
)
heat_data <- full_grid %>%
  left_join(hit_summary,
            by = c("pathway", "pathway_group", "Sample_Source", "Environment")) %>%
  mutate(prevalence  = replace_na(prevalence, 0),
         mean_genes  = replace_na(mean_genes, 0),
         n_mags_with = replace_na(n_mags_with, 0L)) %>%
  left_join(n_mags_per_source,
            by = c("Sample_Source", "Environment"), suffix = c("", ".y")) %>%
  mutate(n_total = coalesce(n_total, n_total.y)) %>% select(-n_total.y) %>%
  mutate(
    pathway = factor(pathway, levels = rev(pathway_order)),
    pathway_group = factor(pathway_group, levels = group_levels),
    Sample_Source = factor(Sample_Source,
                           levels = c("Abijata", "Chitu", "Shala",
                                      "Goat", "Cattle", "Sheep")),
    prev_label = ifelse(prevalence > 0, sprintf("%.0f", prevalence), "")
  )

# ── Fisher's exact enrichment ────────────────────────────────────────────────
# Start from full pathway list to include 0-hit pathways (Dsr, Cellulosome)
enrich_counts <- hit_raw %>%
  distinct(pathway, pathway_group, MAG, Environment) %>%
  count(pathway, pathway_group, Environment)

enrich_pathway <- pathway_genes %>%
  select(pathway, pathway_group) %>%
  left_join(
    enrich_counts %>% pivot_wider(names_from = Environment, values_from = n, values_fill = 0L),
    by = c("pathway", "pathway_group")
  ) %>%
  mutate(across(any_of(c("Soda Lake", "Rumen")), ~ replace_na(., 0L))) %>%
  rename(SL_pos = `Soda Lake`, RU_pos = Rumen) %>%
  mutate(SL_neg = n_sl_mags - SL_pos, RU_neg = n_ru_mags - RU_pos) %>%
  # Skip Fisher's test for pathways with 0 hits in both environments
  filter(SL_pos + RU_pos > 0) %>%
  rowwise() %>%
  mutate(
    ft = list(fisher.test(matrix(c(SL_pos, SL_neg, RU_pos, RU_neg), nrow = 2))),
    fisher_p   = ft$p.value,
    odds_ratio = ft$estimate
  ) %>% ungroup() %>% select(-ft) %>%
  mutate(
    p_adj = p.adjust(fisher_p, method = "BH"),
    log2_OR = log2(odds_ratio + 0.01),
    enriched_in = ifelse(log2_OR > 0, "Soda Lake", "Rumen"),
    sig_label = case_when(
      p_adj < 0.001 ~ "***", p_adj < 0.01 ~ "**",
      p_adj < 0.05  ~ "*",   TRUE ~ "ns"
    ),
    pathway = factor(pathway, levels = rev(pathway_order)),
    pathway_group = factor(pathway_group, levels = group_levels)
  )

# ══════════════════════════════════════════════════════════════════════════════
# Pathway heatmap with abbreviated strip labels + inline significance
# ══════════════════════════════════════════════════════════════════════════════

# Abbreviate pathway group labels to avoid overlap
group_abbrev <- c("pH homeostasis" = "pH", "Oxidative stress" = "Oxidative",
                  "Osmotic stress" = "Osmotic", "Sulfur cycling" = "Sulfur",
                  "Carbon fixation" = "C fixation", "Nitrogen cycling" = "Nitrogen",
                  "Carbohydrate" = "Carbohydrate", "Fermentation" = "Fermentation",
                  "Metal resistance" = "Metal")
heat_data <- heat_data %>%
  mutate(pathway_group_short = factor(group_abbrev[as.character(pathway_group)],
                                       levels = group_abbrev))

# Add enrichment significance annotations to heatmap data (right-side label)
enrich_annot <- enrich_pathway %>%
  mutate(
    direction = ifelse(log2_OR > 0, "\u25B2 SL", "\u25BC RU"),
    enrich_label = ifelse(sig_label == "ns", "ns",
                          paste0(sig_label, " ", direction))
  ) %>%
  select(pathway, enrich_label, sig_label, enriched_in)

heat_data <- heat_data %>%
  left_join(enrich_annot, by = "pathway")

p_fig4 <- ggplot(heat_data, aes(x = Sample_Source, y = pathway, fill = prevalence)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = prev_label),
            size = 1.8, colour = "grey20", fontface = "plain") +
  facet_grid(pathway_group_short ~ Environment, scales = "free", space = "free",
             switch = "y", drop = TRUE) +
  scale_x_discrete(drop = TRUE) +
  scale_fill_gradientn(
    name = "Prevalence (% MAGs)",
    colours = c("#F7F7F7", "#D1E5F0", "#92C5DE", "#4393C3",
                "#FDDBC7", "#F4A582", "#D6604D", "#B2182B"),
    values  = scales::rescale(c(0, 5, 15, 30, 40, 55, 75, 100)),
    limits  = c(0, 100),
    breaks  = c(0, 10, 25, 50, 75, 100),
    na.value = "#F7F7F7"
  ) +
  labs(x = NULL, y = NULL,
       caption = "Fisher\u2019s exact test (BH-adjusted): * p < 0.05, ** p < 0.01, *** p < 0.001") +
  theme_pub(base_size = 8) +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1, size = 6.5),
    axis.text.y    = element_text(size = 6),
    axis.line      = element_blank(),
    axis.ticks     = element_blank(),
    strip.text.y.left = element_text(angle = 0, face = "bold", size = 6),
    strip.text.x      = element_text(face = "bold", size = 7),
    strip.placement    = "outside",
    strip.background   = element_rect(fill = "grey95", colour = NA),
    panel.spacing.x    = unit(0.5, "lines"),
    panel.spacing.y    = unit(0.3, "lines"),
    panel.background   = element_rect(fill = "#FCFCFC", colour = NA),
    legend.position    = "bottom",
    legend.key.height  = unit(0.25, "cm"),
    legend.key.width   = unit(1.5, "cm"),
    legend.title       = element_text(size = 7, face = "bold"),
    legend.text        = element_text(size = 6),
    plot.caption       = element_text(size = 6, colour = "grey50", hjust = 1)
  )
```

## 4.4 Panel D — GH Family × Genus Heatmap

```{r fig3-panelD-gh-heatmap, fig.height=7, fig.width=12.60}
# ── Build GH genus contribution matrix ───────────────────────────────────────
gh_genus <- cazyme_2t %>%
  filter(CAZyme_Class == "GH", !is.na(CAZyme_Family)) %>%
  left_join(
    novelty_rank %>% select(user_genome, g, p) %>%
      rename(Bin_ID = user_genome, Genus = g, Phylum = p),
    by = "Bin_ID"
  ) %>%
  mutate(Genus = ifelse(is.na(Genus) | Genus == "", paste0("Novel (", Phylum, ")"), Genus))

# Focus on key GH families
key_gh <- c("GH1", "GH2", "GH3", "GH5", "GH9", "GH10", "GH13", "GH18", "GH23", "GH26", "GH28", "GH31", "GH43", "GH103")

# Top contributing genera per environment (top 10 overall to keep heatmap compact)
top_genera <- gh_genus %>%
  filter(CAZyme_Family %in% key_gh) %>%
  count(Genus, Environment, name = "total") %>%
  group_by(Environment) %>%
  slice_max(total, n = 10, with_ties = FALSE) %>%
  pull(Genus) %>% unique()

# Gene count matrix: Genus × GH family, split by environment
gh_mat_data <- gh_genus %>%
  filter(CAZyme_Family %in% key_gh, Genus %in% top_genera) %>%
  count(Genus, CAZyme_Family, Environment)

# Build separate matrices for SL and Rumen, then combine
gh_wide <- gh_mat_data %>%
  pivot_wider(names_from = CAZyme_Family, values_from = n, values_fill = 0L)

# Ensure all key GH families appear
for (fam in key_gh) {
  if (!fam %in% colnames(gh_wide)) gh_wide[[fam]] <- 0L
}

# Sort GH families numerically
gh_fam_sorted <- key_gh[order(as.numeric(str_extract(key_gh, "\\d+")))]

mat_gh <- gh_wide %>%
  arrange(Environment, desc(rowSums(across(all_of(gh_fam_sorted))))) %>%
  as.data.frame()
rownames(mat_gh) <- paste0(mat_gh$Genus, " (", mat_gh$Environment, ")")
env_vec <- mat_gh$Environment
mat_gh_num <- as.matrix(mat_gh[, gh_fam_sorted])

# ── Row annotation: environment ──────────────────────────────────────────────
ha_gh_row <- rowAnnotation(
  Environment = env_vec,
  col = list(Environment = env_colors),
  annotation_name_gp = gpar(fontsize = 7, fontfamily = pub_font),
  simple_anno_size = unit(3, "mm"),
  annotation_legend_param = list(
    title_gp = gpar(fontsize = 7, fontface = "bold", fontfamily = pub_font),
    labels_gp = gpar(fontsize = 6, fontfamily = pub_font)
  )
)

# ── Heatmap ──────────────────────────────────────────────────────────────────
# Use non-linear breakpoints to improve visual contrast in the 0-50 range
# where most data clusters (linear scale washes out low-count differences)
max_val <- max(mat_gh_num)
col_fun_gh <- colorRamp2(c(0, 5, 20, 50, 100, max_val),
                          c("#F7F7F7", "#D1E5F0", "#92C5DE",
                            "#4393C3", "#B2182B", "#67001F"))

ht_gh_genus <- Heatmap(
  mat_gh_num,
  name = "Gene count",
  col = col_fun_gh,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = env_vec,
  row_gap = unit(2, "mm"),
  row_title_gp = gpar(fontsize = 8, fontface = "bold", fontfamily = pub_font),
  row_names_gp = gpar(fontsize = 6, fontface = "italic", fontfamily = pub_font),
  column_names_gp = gpar(fontsize = 7, fontface = "bold", fontfamily = pub_font),
  column_names_rot = 45,
  left_annotation = ha_gh_row,
  border = TRUE,
  rect_gp = gpar(col = "white", lwd = 0.4),
  cell_fun = function(j, i, x, y, width, height, fill) {
    val <- mat_gh_num[i, j]
    if (val > 0) {
      grid.text(val, x, y, gp = gpar(fontsize = 5, fontfamily = pub_font,
                                       col = ifelse(val > 40, "white", "grey20")))
    }
  },
  column_title = "Top genera contributing to key GH families",
  column_title_gp = gpar(fontsize = 9, fontface = "bold", fontfamily = pub_font),
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 7, fontface = "bold", fontfamily = pub_font),
    labels_gp = gpar(fontsize = 6, fontfamily = pub_font)
  )
)
draw(ht_gh_genus)
save_complexheatmap(ht_gh_genus, "Fig6_gh_genus_heatmap", nm_wide, nm_tall * 0.6)
```

## 4.5 ★ Figure 4: Functional Pathway Heatmap (standalone)

```{r fig4-pathway-standalone, fig.height=12, fig.width=12.60}
# Figure 4: Functional pathway heatmap — standalone full-width figure
p_fig4_path <- p_fig4
p_fig4_path
save_fig(p_fig4_path, "Fig4_functional_pathways", nm_wide, 12)
```

## 4.6 ★ Figure 5: CAZyme Diversity & Capacity

```{r fig5-cazyme-composite, fig.height=14, fig.width=12.60}
# Panel A — CAZyme richness vs completeness QC
# Panel B — CAZyme class distribution
# Panel C — Per-MAG CAZyme richness, D — Shannon diversity
# Panel E — Sequence identity density

p_fig5_caz <- (p_caz_qual + labs(title = NULL, subtitle = NULL, tag = "A")) /
               (p_caz_class + labs(title = NULL, tag = "B")) /
               ((p_r + labs(title = NULL, tag = "C")) |
                (p_s + labs(title = NULL, tag = "D"))) /
               (p_caz_id + labs(title = NULL, subtitle = NULL, tag = "E")) +
  plot_layout(heights = c(0.9, 1, 1, 1)) &
  theme_pub() &
  theme(plot.tag = element_text(size = 10, face = "bold"),
        plot.margin = margin(4, 4, 4, 4))
p_fig5_caz
save_fig(p_fig5_caz, "Fig5_cazyme_diversity", nm_wide, 14)
```

## 4.7 ★ Figure 6: GH Family × Genus Repertoire

```{r fig6-gh-heatmap-standalone, fig.height=9, fig.width=12.60}
# GH family × genus heatmap as standalone Figure 6 (15 GH families)
p_fig6_gh <- wrap_elements(full = grid::grid.grabExpr(
  draw(ht_gh_genus, padding = unit(c(2, 2, 2, 2), "mm"))
))
p_fig6_gh
save_fig(p_fig6_gh, "Fig6_gh_genus_heatmap", nm_wide, 9)
```

---

## Supplementary Figure S3 — CAZyme NMDS + GH Volcano

```{r figS3-nmds-volcano, fig.height=6, fig.width=12.60}
# NMDS ordination of CAZyme profiles
set.seed(42)
nmds_caz <- metaMDS(fam_matrix, distance = "bray", k = 2, trymax = 100, trace = FALSE)
perm_caz <- adonis2(fam_matrix ~ Environment, data = mag_cazyme, method = "bray", permutations = 999)

nmds_caz_df <- as_tibble(scores(nmds_caz, display = "sites")) %>%
  mutate(Bin_ID = mag_cazyme$Bin_ID, Environment = mag_cazyme$Environment,
         Sample_Source = mag_cazyme$Sample_Source)

p_caz_nmds <- ggplot(nmds_caz_df, aes(x = NMDS1, y = NMDS2, colour = Sample_Source,
                                        shape = Environment)) +
  geom_point(size = 1.8, alpha = 0.7) +
  stat_ellipse(aes(group = Environment), linetype = "dashed", linewidth = 0.5, colour = "grey40") +
  scale_colour_manual(values = source_colors) +
  scale_shape_manual(values = c("Soda Lake" = 16, "Rumen" = 17)) +
  labs(subtitle = sprintf("Stress = %.3f | PERMANOVA R² = %.3f, p = %.3f",
                           nmds_caz$stress, perm_caz$R2[1], perm_caz$`Pr(>F)`[1]),
       colour = "Source", shape = "Environment") +
  theme_pub()

# GH family enrichment volcano
gh_data <- cazyme_2t %>%
  filter(CAZyme_Class == "GH", !is.na(CAZyme_Family)) %>%
  mutate(Sample = str_extract(Bin_ID, "^[A-Za-z]+\\d*"))

gh_enrich <- gh_data %>%
  count(Environment, CAZyme_Family) %>%
  pivot_wider(names_from = Environment, values_from = n, values_fill = 0)

total_soda  <- sum(gh_data$Environment == "Soda Lake")
total_rumen <- sum(gh_data$Environment == "Rumen")

gh_fisher <- gh_enrich %>%
  rowwise() %>%
  mutate(
    fisher_p = fisher.test(matrix(c(`Soda Lake`, total_soda - `Soda Lake`,
                                     Rumen, total_rumen - Rumen), nrow = 2))$p.value,
    odds_ratio = fisher.test(matrix(c(`Soda Lake`, total_soda - `Soda Lake`,
                                       Rumen, total_rumen - Rumen), nrow = 2))$estimate,
    log2_OR = log2(odds_ratio + 0.01)
  ) %>% ungroup() %>%
  mutate(
    p_adj = p.adjust(fisher_p, method = "BH"),
    significant = p_adj < 0.05,
    enriched_in = case_when(!significant ~ "NS", log2_OR > 0 ~ "Soda Lake",
                            log2_OR < 0 ~ "Rumen", TRUE ~ "NS")
  ) %>% arrange(p_adj)

p_gh_volcano <- ggplot(gh_fisher, aes(x = log2_OR, y = -log10(p_adj),
                                       colour = enriched_in, size = `Soda Lake` + Rumen)) +
  geom_point(alpha = 0.65) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50", linewidth = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.3) +
  geom_text_repel(
    data = gh_fisher %>% filter(significant) %>% slice_min(p_adj, n = 15),
    aes(label = CAZyme_Family), size = 2.2, max.overlaps = 25
  ) +
  scale_colour_manual(values = c("Soda Lake" = npg_cols[1], "Rumen" = npg_cols[4], "NS" = "grey75")) +
  scale_size_continuous(range = c(1, 5), name = "Total genes") +
  labs(x = "log2(Odds Ratio)", y = "-log10(adjusted p-value)", colour = "Enriched in") +
  theme_pub()

p_figS3 <- (p_caz_nmds + labs(title = NULL)) | (p_gh_volcano + labs(title = NULL))
p_figS3
save_fig(p_figS3, "FigS3_nmds_volcano", nm_wide, nm_tall * 0.5)
```

---

# 5 — Functional Capacity (Prokka) — Reference 
## 5.1 Define Functional Gene Categories

```{r prokka-gene-categories}
func_categories <- tribble(
  ~Category,                    ~gene_pattern,                                       ~label,
  # pH homeostasis
  "pH homeostasis",             "^(mrpA|mrpB|mrpC|mrpD)$",                           "Mrp (Na+/H+ antiporter complex)",
  "pH homeostasis",             "^(nhaA|nhaC)$",                                     "NhaA/C (Na+/H+ antiporter)",
  # Oxidative stress
  "Oxidative stress",           "^(katG|katA)$",                                     "Catalase / catalase-peroxidase",
  "Oxidative stress",           "^(sodA|sodB)$",                                     "Superoxide dismutase",
  "Oxidative stress",           "^(ahpC)$",                                          "Alkyl hydroperoxide reductase",
  "Oxidative stress",           "^(cspA|cspB)$",                                     "Cold shock protein",
  # Osmotic stress
  "Osmotic stress",             "^(ectA|ectB|ectC)$",                                "ect (ectoine biosynthesis)",
  "Osmotic stress",             "^(betA|betB|betT)$",                                "bet (glycine betaine)",
  "Osmotic stress",             "^(proV|proW|proX)$",                                "pro (proline/betaine transport)",
  "Osmotic stress",             "^(kdpA|kdpB)$",                                     "Kdp (K+ ATPase)",
  # Sulfur cycling
  "Sulfur cycling",             "^(soxA|soxB|soxC|soxY)$",                           "sox (thiosulfate oxidation)",
  "Sulfur cycling",             "^(aprA|aprB)$",                                     "apr (sulfite oxidation)",
  "Sulfur cycling",             "^(dsrA|dsrB|dsrC)$",                                "dsr (dissimilatory sulfate)",
  # Carbon fixation
  "Carbon fixation",            "^(cbbL|cbbS|cbbM|rbcL|rbcS)$",                      "RuBisCO (carbon fixation)",
  # Nitrogen cycling
  "Nitrogen cycling",           "^(nifH|nifD|nifK)$",                                "nif (nitrogen fixation)",
  "Nitrogen cycling",           "^(nirK|nirS)$",                                     "nir (nitrite reduction)",
  "Nitrogen cycling",           "^(narG|narH|narI)$",                                "nar (nitrate reduction)",
  "Nitrogen cycling",           "^(nosZ)$",                                          "nosZ (N2O reduction)",
  "Nitrogen cycling",           "^(norB|norC)$",                                     "nor (NO reduction)",
  "Nitrogen cycling",           "^(ureA|ureB|ureC)$",                                "Urease (urea hydrolysis)",
  # Carbohydrate catabolism
  "Carbohydrate",               "^(cipA|cipB|cipC)$",                                "Cellulosome scaffoldin",
  "Carbohydrate",               "^(susC|susD)$",                                     "SusCD (polysaccharide utilization)",
  "Carbohydrate",               "^(celA|celB)$",                                     "Endoglucanase",
  "Carbohydrate",               "^(bglA|bglB)$",                                     "Beta-glucosidase",
  "Carbohydrate",               "^(xylA|xylB)$",                                     "Xylose utilization",
  "Carbohydrate",               "^(glgP)$",                                          "Glycogen phosphorylase",
  # Fermentation
  "Fermentation",               "^(ackA|pta)$",                                      "Acetate production",
  "Fermentation",               "^(pflB)$",                                          "Formate acetyltransferase",
  "Fermentation",               "^(pccB)$",                                          "Propionyl-CoA carboxylase",
  # Metal resistance
  "Metal resistance",           "^(arsC)$",                                          "Arsenate reductase"
)

func_product_patterns <- tribble(
  ~Category,              ~product_pattern,
  "pH homeostasis",       "(?i)(Na\\+/H\\+|sodium.*proton|mrp.*antiporter|nha.*antiporter)",
  "Oxidative stress",     "(?i)(catalase|superoxide dismutase|peroxide reductase|cold shock protein)",
  "Osmotic stress",       "(?i)(ectoine|compatible solute|glycine betaine|choline dehydrogenase|potassium.*ATPase)",
  "Sulfur cycling",       "(?i)(thiosulfate|sulfite|sulfur oxidat|sulfate reduc)",
  "Carbon fixation",      "(?i)(ribulose.*bisphosphate|RuBisCO|carbon fixation)",
  "Nitrogen cycling",     "(?i)(nitrogenase|nitrate reduc|nitrite reduc|nitrogen fix|nitrous oxide|urease)",
  "Carbohydrate",         "(?i)(endoglucanase|beta-glucosidase|xylanase|xylose isomerase|cellulosome|scaffoldin|pullulanase|glycogen phosphorylase|SusC|SusD)",
  "Fermentation",         "(?i)(acetate kinase|phosphotransacetylase|formate acetyltransferase|propionyl-CoA carboxylase)",
  "Metal resistance",     "(?i)(arsenate reductase|arsenical)"
)
```

## 5.2 Gene-level Functional Profiling

```{r prokka-func-count}
gene_hits <- prokka_ann %>%
  filter(!is.na(gene), gene != "") %>%
  # Remove known false-positive gene-name collisions (same as heatmap pipeline)
  filter(!(gene %in% c("soxA","soxB","soxD") & str_detect(product, regex("sarcosine", ignore_case = TRUE))),
         !(gene == "proX" & str_detect(product, regex("prolyl-tRNA", ignore_case = TRUE)))) %>%
  rowwise() %>%
  mutate(
    matched_cat = {
      cats <- func_categories %>% filter(str_detect(gene, gene_pattern)) %>% pull(Category) %>% unique()
      if (length(cats) == 0) NA_character_ else paste(cats, collapse = "; ")
    },
    matched_label = {
      labs <- func_categories %>% filter(str_detect(gene, gene_pattern)) %>% pull(label) %>% unique()
      if (length(labs) == 0) NA_character_ else paste(labs, collapse = "; ")
    }
  ) %>% ungroup() %>% filter(!is.na(matched_cat))

product_hits <- prokka_ann %>%
  filter(!is.na(product), product != "hypothetical protein") %>%
  rowwise() %>%
  mutate(
    matched_cat = {
      cats <- func_product_patterns %>% filter(str_detect(product, product_pattern)) %>% pull(Category) %>% unique()
      if (length(cats) == 0) NA_character_ else paste(cats, collapse = "; ")
    }
  ) %>% ungroup() %>% filter(!is.na(matched_cat))

gene_func <- gene_hits %>%
  select(MAG, locus_tag, gene, product, Environment, Sample_Source,
         Category = matched_cat, Label = matched_label)

product_func <- product_hits %>%
  anti_join(gene_hits, by = "locus_tag") %>%
  select(MAG, locus_tag, gene, product, Environment, Sample_Source, Category = matched_cat) %>%
  mutate(Label = Category)

all_func <- bind_rows(gene_func, product_func)
func_long <- all_func %>% separate_rows(Category, sep = "; ")
```


## 5.3 Functional Pathway Heatmap (computed in Section 4.4)

_The pathway heatmap (`p_fig4`) was computed in section 4.4 and is incorporated into Figure 3 (panel A). The standalone pathway figure is retained below for reference._

```{r fig4-reference, fig.width=12.60, fig.height=7.5}
# p_fig4 already built in section 4.4
p_fig4
```

---

# 6 — Phylogenomic Tree — Figure 6 
## 6.1 Prepare Tree Metadata

```{r tree-metadata}
bb_mag_tips <- tree_backbone$tip.label[
  !str_detect(tree_backbone$tip.label, "^(GB_|RS_|GCA_|GCF_)")
]
cat(sprintf("MAGs in backbone tree: %d / %d total\n", length(bb_mag_tips), nrow(red_df)))

tree_mag_only <- drop.tip(tree_backbone, setdiff(tree_backbone$tip.label, bb_mag_tips))

tree_meta_full <- tibble(label = tree_mag_only$tip.label) %>%
  left_join(
    novelty_rank %>%
      select(user_genome, p, c, o, f, g, Environment) %>%
      rename(label = user_genome, Phylum = p, Class = c, Order = o, Family = f, Genus = g),
    by = "label"
  ) %>%
  mutate(
    Environment = ifelse(is.na(Environment), assign_env(label), Environment),
    Source = assign_source(label),
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", Phylum),
    Genus  = ifelse(is.na(Genus) | Genus == "", "Novel", Genus)
  )

top10_phy_tree <- tree_meta_full %>%
  filter(Phylum != "Unclassified") %>%
  count(Phylum, sort = TRUE) %>% slice_head(n = 10) %>% pull(Phylum)

tree_meta_full <- tree_meta_full %>%
  mutate(Phylum_plot = ifelse(Phylum %in% top10_phy_tree, Phylum, "Other"),
         # Keep Environment as character for ggtree — gheatmap() coerces
         # factors to integer codes which breaks scale_shape_manual
         Environment = as.character(Environment))

phy_tree_pal <- c(setNames(pal_d3("category10")(10), sort(top10_phy_tree)),
                  "Other" = "grey65", "Unclassified" = "grey85")
```

## 6.2 ★ Figure 6: Circular Phylogenomic Tree

```{r fig5-hero, fig.height=18, fig.width=18}
p_base <- ggtree(tree_mag_only, layout = "circular", size = 0.15,
                  colour = "grey50") %<+% tree_meta_full

p_tree_ann <- p_base +
  geom_tippoint(aes(colour = Source, shape = Environment), size = 1.8, alpha = 0.85) +
  scale_colour_manual(values = source_colors, na.value = "grey80", name = "Sample Source") +
  scale_shape_manual(values = c("Soda Lake" = 16, "Rumen" = 17), name = "Environment")

ring_df <- tree_meta_full %>% select(label, Phylum_plot)

p_tree_hero <- gheatmap(p_tree_ann, ring_df %>% column_to_rownames("label"),
                         offset = 0.02, width = 0.08, colnames = FALSE, color = NA) +
  scale_fill_manual(values = phy_tree_pal, name = "Phylum",
                    guide = guide_legend(order = 1)) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 9, family = pub_font),
    legend.position = "right",
    legend.text = element_text(size = 6, family = pub_font),
    legend.title = element_text(size = 7, face = "bold", family = pub_font),
    legend.key.size = unit(0.3, "cm"),
    legend.spacing.y = unit(0.15, "cm")
  ) +
  guides(
    fill = guide_legend(order = 1, ncol = 1, override.aes = list(size = 3)),
    colour = guide_legend(order = 2, override.aes = list(size = 2.5)),
    shape = guide_legend(order = 3, override.aes = list(size = 2.5))
  )
p_tree_hero

save_fig(p_tree_hero, "Fig3_phylogenomic_tree", nm_wide, nm_wide)
```

---

# 7 — Structural Analysis (AlphaFold3 + Foldseek) — Figure 7 
## 7.1 Load Structural Data

```{r struct-load}
# ── Foldseek best hits (v2: 12 structures, 6 GH families) ────────────────────
foldseek_best <- read_tsv(file.path(data_dir, "foldseek_best_hits.tsv"), show_col_types = FALSE)

# ── AlphaFold3 summary (pTM, pLDDT from model_0 JSONs) ──────────────────────
af3_summary <- read_tsv(file.path(data_dir, "alphafold3_summary.tsv"), show_col_types = FALSE)

# ── Merge Foldseek hits with AF3 summary ─────────────────────────────────────
# Query names in foldseek_best_hits.tsv match Structure names in af3_summary
# (e.g. "soda_gh1", "rumen_gh10")
struct_df <- foldseek_best %>%
  rename(Structure = Query) %>%
  left_join(af3_summary, by = "Structure") %>%
  mutate(
    Source = Environment,
    GH_Family = str_extract(Structure, "gh\\d+(_\\d+)?") %>%
      str_to_upper() %>% str_replace("_", " sub"),
    GH_Family_short = str_extract(Structure, "gh\\d+") %>% str_to_upper(),
    PDB_ID = str_extract(Best_PDB_Hit, "^[a-z0-9]+")
  )

gh_info <- tribble(
  ~GH_Family_short, ~Substrate,               ~Mechanism,
  "GH1",           "beta-glucoside",          "Retaining",
  "GH3",           "beta-glucoside",          "Retaining",
  "GH5",           "Cellulose",               "Retaining",
  "GH9",           "Cellulose",               "Inverting",
  "GH10",          "Xylan",                   "Retaining",
  "GH28",          "Pectin / galacturonan",   "Inverting"
)

struct_df <- struct_df %>% left_join(gh_info, by = "GH_Family_short")

# ── Active site conservation data ────────────────────────────────────────────
# Generated by the Python processing pipeline (from AlphaFold3 CIF files)
active_site_file <- file.path("renders", "active_site_conservation.tsv")
active_sites <- read_tsv(active_site_file, show_col_types = FALSE)
cat(sprintf("Loaded %d active site annotations from %d structures\n",
            nrow(active_sites), n_distinct(active_sites$Structure)))

# ── Per-residue pLDDT profiles ───────────────────────────────────────────────
plddt_df <- read_tsv(file.path(data_dir, "plddt_profiles.tsv"), show_col_types = FALSE)
cat(sprintf("Loaded pLDDT profiles: %d residues across %d structures\n",
            nrow(plddt_df), n_distinct(plddt_df$Structure)))

# ── (AF3 summary already loaded and merged in struct_df above) ───────────────
```

## 7.2 Structure Renders (PyMOL)

```{r fig6-renders, fig.height=21, fig.width=16}
# ── Load PyMOL renders ────────────────────────────────────────────────────────
# Generated by: pymol scripts/run_pymol_renders.py
# Main figure uses pLDDT-coloured individual structures
# Superposition + close-up go to Extended Data (Supplementary Figure S3)

families <- c("GH1", "GH3", "GH5_11", "GH9", "GH10", "GH28")

# Short labels for panel captions (env + family + function)
panel_labels <- c(
  "soda_GH1"    = "GH1 Soda Lake",     "rumen_GH1"    = "GH1 Rumen",
  "soda_GH3"    = "GH3 Soda Lake",     "rumen_GH3"    = "GH3 Rumen",
  "soda_GH5_11" = "GH5_11 Soda Lake",  "rumen_GH5_11" = "GH5_11 Rumen",
  "soda_GH9"    = "GH9 Soda Lake",     "rumen_GH9"    = "GH9 Rumen",
  "soda_GH10"   = "GH10 Soda Lake",    "rumen_GH10"   = "GH10 Rumen",
  "soda_GH28"   = "GH28 Soda Lake",    "rumen_GH28"   = "GH28 Rumen"
)

fam_to_key <- c("GH1" = "gh1", "GH3" = "gh3", "GH5_11" = "gh5_11",
                "GH9" = "gh9", "GH10" = "gh10", "GH28" = "gh28")

load_render <- function(env_prefix, fam_key, fallback_label = "") {
  img_path <- paste0("figure7_", env_prefix, "_", fam_key, ".png")
  if (!file.exists(img_path)) {
    fam_upper <- toupper(fam_key) %>% str_replace("gh", "GH")
    suffix <- ifelse(env_prefix == "soda", "_SL_plddt", "_RU_plddt")
    img_path <- file.path("renders", paste0("Fig6_", fam_upper, suffix, ".png"))
  }
  if (file.exists(img_path)) {
    img <- png::readPNG(img_path)
    wrap_elements(full = grid::rasterGrob(img, interpolate = TRUE))
  } else {
    ggplot() +
      annotate("text", x = 0.5, y = 0.5,
               label = paste0(fam_key, " ", fallback_label, "\n(run PyMOL to generate)"),
               size = 3, colour = "grey50", hjust = 0.5) +
      theme_void() +
      theme(plot.background = element_rect(fill = "grey97", colour = "grey80"))
  }
}

# ── MAIN FIGURE 7: 6×2 grid layout ──────────────────────────────────────────
# Columns: Soda Lake | Rumen
# Rows:    GH1, GH3, GH5_11, GH9, GH10, GH28
# Panel order (row-wise, left-to-right):
#   A=GH1-SL  B=GH1-RU
#   C=GH3-SL  D=GH3-RU
#   E=GH5_11-SL  F=GH5_11-RU
#   G=GH9-SL  H=GH9-RU
#   I=GH10-SL J=GH10-RU
#   K=GH28-SL L=GH28-RU

panels <- list()
# Row 1 — GH1
panels[["A"]] <- load_render("soda",  "gh1")
panels[["B"]] <- load_render("rumen", "gh1")
# Row 2 — GH3
panels[["C"]] <- load_render("soda",  "gh3")
panels[["D"]] <- load_render("rumen", "gh3")
# Row 3 — GH5_11
panels[["E"]] <- load_render("soda",  "gh5_11")
panels[["F"]] <- load_render("rumen", "gh5_11")
# Row 4 — GH9
panels[["G"]] <- load_render("soda",  "gh9")
panels[["H"]] <- load_render("rumen", "gh9")
# Row 5 — GH10
panels[["I"]] <- load_render("soda",  "gh10")
panels[["J"]] <- load_render("rumen", "gh10")
# Row 6 — GH28
panels[["K"]] <- load_render("soda",  "gh28")
panels[["L"]] <- load_render("rumen", "gh28")

# Column headers (environment) and row labels (GH family)
col_heads <- c("Soda Lake", "Rumen")

# Labels for supplementary superposition panels
family_labels <- c(
  "GH1"    = "GH1  \u03b2-glucosidase",
  "GH3"    = "GH3  \u03b2-glucosidase",
  "GH5_11" = "GH5\u2081\u2081  Endoglucanase",
  "GH9"    = "GH9  Endoglucanase",
  "GH10"   = "GH10  Xylanase",
  "GH28"   = "GH28  Polygalacturonase"
)

p_renders <- (
  # Row 1 — GH1:    Soda Lake | Rumen
  (panels[["A"]] | panels[["B"]]) /
  # Row 2 — GH3:    Soda Lake | Rumen
  (panels[["C"]] | panels[["D"]]) /
  # Row 3 — GH5_11: Soda Lake | Rumen
  (panels[["E"]] | panels[["F"]]) /
  # Row 4 — GH9:    Soda Lake | Rumen
  (panels[["G"]] | panels[["H"]]) /
  # Row 5 — GH10:   Soda Lake | Rumen
  (panels[["I"]] | panels[["J"]]) /
  # Row 6 — GH28:   Soda Lake | Rumen
  (panels[["K"]] | panels[["L"]])
) +
  plot_layout(heights = c(1, 1, 1, 1, 1, 1)) +
  plot_annotation(tag_levels = "A") &
  # Inset the panel tag into the top-left corner of each panel so the
  # label doesn't reserve a whole header strip above each structure.
  # Negative margins pull adjacent panels tight against each other.
  theme(plot.tag = element_text(size = 14, face = "bold", family = pub_font,
                                  hjust = 0, vjust = 1),
        plot.tag.position = c(0.02, 0.98),
        plot.margin = margin(0, 0, 0, 0))
p_renders

# ── SUPPLEMENTARY: Superposition + Close-up (for Supplementary Fig S3) ────────
sup_rows <- list()
for (fam in families) {
  p_sup  <- load_render(fam, "_superpose", "overlay")
  p_zoom <- load_render(fam, "_closeup", "active site")

  sup_rows[[fam]] <- (p_sup | p_zoom) +
    plot_layout(widths = c(1, 1)) +
    plot_annotation(subtitle = family_labels[[fam]]) &
    theme(plot.subtitle = element_text(face = "bold", size = 9, hjust = 0,
                                       family = pub_font))
}

p_superpositions <- sup_rows[["GH1"]] / sup_rows[["GH3"]] /
                    sup_rows[["GH5_11"]] / sup_rows[["GH9"]] /
                    sup_rows[["GH10"]] / sup_rows[["GH28"]] +
  plot_layout(heights = c(1, 1, 1, 1, 1, 1)) +
  plot_annotation(
    title = "Structural superposition and active-site conservation",
    caption = "Left: overlay (Soda Lake = red, Rumen = teal).  Right: active-site close-up with catalytic residues labelled."
  ) &
  theme(plot.title = element_text(size = 10, face = "bold", family = pub_font),
        plot.caption = element_text(size = 7, hjust = 0.5, colour = "grey40",
                                     family = pub_font))
```

## 7.3 Panel B — Per-Residue pLDDT Confidence Profiles

```{r fig6-panelB-plddt, fig.height=4, fig.width=12.60}
# ── pLDDT ribbon plot: per-residue confidence coloured by AlphaFold bins ─────
# Mark active site positions with vertical lines

plddt_plot <- plddt_df %>%
  mutate(
    # Derive GH subfamily from structure name
    GH_Subfamily = case_when(
      str_detect(Structure, "gh5_11") ~ "GH5_11",
      TRUE ~ GH_Family
    ),
    confidence_bin = case_when(
      pLDDT > 90 ~ "Very high (>90)",
      pLDDT > 70 ~ "Confident (70-90)",
      pLDDT > 50 ~ "Low (50-70)",
      TRUE ~ "Very low (<50)"
    ),
    confidence_bin = factor(confidence_bin,
      levels = c("Very high (>90)", "Confident (70-90)", "Low (50-70)", "Very low (<50)")),
    # Short label for facet
    short_label = paste0(GH_Subfamily, " (", ifelse(Environment == "Soda Lake", "SL", "RU"), ")")
  )

# Active site positions for annotation
active_marks <- active_sites %>%
  mutate(
    GH_Subfamily = case_when(
      str_detect(Structure, "gh5_11") ~ "GH5_11",
      TRUE ~ GH_Family
    ),
    short_label = paste0(GH_Subfamily, " (",
                         ifelse(Environment == "Soda Lake", "SL", "RU"), ")"),
    mark_label = paste0(Pred_Residue, Pred_Position)
  )

plddt_colors <- c(
  "Very high (>90)" = "#0053D6",
  "Confident (70-90)" = "#65CBF3",
  "Low (50-70)" = "#FFDB13",
  "Very low (<50)" = "#FF7D45"
)

p_plddt <- ggplot(plddt_plot, aes(x = Residue_Position, y = pLDDT)) +
  geom_line(aes(colour = confidence_bin), linewidth = 0.4, show.legend = FALSE) +
  geom_point(aes(colour = confidence_bin), size = 0.1, alpha = 0.4) +
  # Active site markers
  geom_vline(data = active_marks, aes(xintercept = Pred_Position),
             linetype = "dashed", colour = "red", linewidth = 0.4, alpha = 0.7) +
  geom_text(data = active_marks,
            aes(x = Pred_Position, y = 102, label = mark_label),
            size = 1.8, angle = 90, hjust = 0, colour = "red") +
  # Confidence thresholds
  geom_hline(yintercept = 90, linetype = "dotted", colour = "grey60", linewidth = 0.2) +
  geom_hline(yintercept = 70, linetype = "dotted", colour = "grey60", linewidth = 0.2) +
  facet_wrap(~ short_label, scales = "free_x", nrow = 2) +
  scale_colour_manual(values = plddt_colors, name = "Confidence") +
  scale_y_continuous(limits = c(20, 105), breaks = c(50, 70, 90, 100)) +
  labs(x = "Residue position", y = "pLDDT") +
  theme_pub() +
  theme(
    strip.text = element_text(face = "bold", size = 7),
    axis.text.x = element_text(size = 5),
    legend.position = "bottom",
    legend.key.size = unit(0.3, "cm"),
    panel.spacing = unit(0.3, "lines")
  ) +
  guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1)))
p_plddt
```

## 7.4 Panel C — Sequence Identity vs Structural Similarity (Scatter)

```{r fig6-panelB, fig.height=5, fig.width=7}
p_struct_scatter <- ggplot(struct_df,
    aes(x = Seq_Identity * 100, y = LDDT,
        colour = Environment, shape = Environment)) +
  geom_point(size = 3.5, alpha = 0.85) +
  geom_text_repel(aes(label = GH_Family), size = 2.5,
                  max.overlaps = 20, segment.size = 0.3) +
  # Quadrant annotations
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50", linewidth = 0.3) +
  geom_vline(xintercept = 40, linetype = "dashed", colour = "grey50", linewidth = 0.3) +
  annotate("rect", xmin = 0, xmax = 40, ymin = 0.5, ymax = 1,
           fill = "#E64B35", alpha = 0.05) +
  annotate("text", x = 20, y = 0.98,
           label = "Novel sequences\nconserved folds",
           size = 2.2, colour = "#E64B35", fontface = "italic") +
  scale_colour_manual(values = env_colors) +
  scale_shape_manual(values = c("Soda Lake" = 16, "Rumen" = 17)) +
  labs(x = "Sequence Identity to Best PDB Hit (%)",
       y = "Foldseek LDDT") +
  theme_pub() +
  theme(legend.position = "right")
p_struct_scatter
```

## 7.5 Panel D — Active Site Conservation

```{r fig6-panelC-activesite, fig.height=5, fig.width=12.60}
# ── Tile plot: GH family × environment, coloured by conservation ─────────────
active_plot <- active_sites %>%
  mutate(
    Conserved = (Conserved == "True"),
    Pred_Position = as.integer(Pred_Position),
    # Derive GH subfamily from structure name
    GH_Subfamily = case_when(
      str_detect(Structure, "gh5_11") ~ "GH5_11",
      TRUE ~ GH_Family
    ),
    Label = paste0(Ref_Residue, Pred_Position, "\n(", Role, ")"),
    Status = ifelse(Conserved, "Conserved", paste0("Substituted\n(", Pred_Residue, ")")),
    GH_Subfamily = factor(GH_Subfamily, levels = c("GH1", "GH3", "GH5_11", "GH9", "GH10", "GH28"))
  )

# Conservation colours: green = conserved, red = substituted
conserv_colors <- c("Conserved" = "#2CA02C", "Substituted" = "#D62728")

# Compute label for each tile
active_plot <- active_plot %>%
  mutate(
    tile_label = ifelse(Conserved,
                        paste0(Pred_Residue, Pred_Position),
                        paste0(Ref_Residue, "->", Pred_Residue, Pred_Position)),
    conserv_status = ifelse(Conserved, "Conserved", "Substituted")
  )

p_active_site <- ggplot(active_plot,
                        aes(x = Role, y = interaction(GH_Subfamily, Environment, sep = " — "),
                            fill = conserv_status)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = tile_label), size = 2.5, fontface = "bold") +
  scale_fill_manual(values = conserv_colors, name = "Catalytic residue") +
  facet_grid(GH_Subfamily ~ ., scales = "free_y", space = "free_y", switch = "y") +
  labs(x = "Catalytic role", y = NULL,
       caption = "Residue labels: 3-letter code + position in predicted structure. Arrow = substitution from reference.") +
  theme_pub() +
  theme(
    axis.text.y = element_text(size = 7),
    axis.text.x = element_text(size = 7, angle = 30, hjust = 1),
    strip.text.y.left = element_text(angle = 0, face = "bold", size = 7.5),
    strip.placement = "outside",
    strip.background = element_rect(fill = "grey96", colour = NA),
    legend.position = "right",
    plot.caption = element_text(size = 5.5, colour = "grey50")
  )
p_active_site
```

## 7.6 Panel E — Structural Metrics Summary (Compact)

```{r fig6-panelD-metrics, fig.height=4, fig.width=7}
# Compact dot plot: LDDT + pTM per structure, one row per query
metrics_long <- struct_df %>%
  select(GH_Family, Environment, LDDT, pTM) %>%
  pivot_longer(cols = c(LDDT, pTM), names_to = "Metric", values_to = "Value") %>%
  mutate(
    Metric = factor(Metric, levels = c("LDDT", "pTM")),
    Label = paste0(GH_Family, "\n(", Environment, ")")
  )

p_metrics <- ggplot(metrics_long,
                    aes(x = Metric, y = Value, colour = Environment)) +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60", linewidth = 0.3) +
  geom_hline(yintercept = 0.8, linetype = "dotted", colour = "grey60", linewidth = 0.3) +
  annotate("text", x = 3.4, y = 0.81, label = "high\nconfidence", size = 2, colour = "grey50") +
  facet_wrap(~ GH_Family, nrow = 1) +
  scale_colour_manual(values = env_colors) +
  scale_y_continuous(limits = c(0.4, 1), breaks = seq(0.4, 1, 0.2)) +
  labs(x = NULL, y = "Score") +
  theme_pub() +
  theme(strip.text = element_text(face = "bold", size = 7),
        axis.text.x = element_text(angle = 40, hjust = 1, size = 6.5),
        legend.position = "right")
p_metrics
```

## 7.7 Table 1 — Structural Results Summary

```{r table1-structural, results='asis'}
# ── Build Table 1: AlphaFold3 + Foldseek + Active-site conservation ──────────
# This reproduces the manuscript Table 1 from raw data so all values are
# traceable and reproducible.

# ── Build Table 1: Full manuscript table ──────────────────────────────────────
# Combines AlphaFold3 predictions, dbCAN/DIAMOND annotations, active-site
# conservation, and Foldseek PDB100 structural similarity for 12 representative
# CAZymes. Source organism and DIAMOND annotation data are curated from the
# original gene-level outputs (protein_taxonomy_mapping.tsv, all_diamond_merged_env.tsv).

# ── Active-site summary per structure (computed from raw data) ────────────────
active_summary <- active_sites %>%
  group_by(Structure) %>%
  summarise(
    n_sites      = n(),
    n_conserved  = sum(Conserved == "True" | Conserved == TRUE),
    residue_str  = paste(
      ifelse(Conserved == "True" | Conserved == TRUE,
             paste0(Pred_Residue, Pred_Position),
             paste0(Pred_Residue, Pred_Position)),
      collapse = ", "
    ),
    plddt_str = paste(
      ifelse(pLDDT_at_site == "N/A", "N/A",
             sprintf("%.1f", as.numeric(pLDDT_at_site))),
      collapse = ", "
    ),
    .groups = "drop"
  )

# ── Curated annotation metadata for the 12 selected structures ───────────────
# Source organism, phylum, and DIAMOND hit statistics were extracted from the
# original dbCAN/DIAMOND annotation pipeline outputs and GTDB-Tk classifications.
# Foldseek PDB hit descriptions are from manual PDB lookup of the best-hit IDs.
curated_meta <- tibble::tribble(
  ~Structure,      ~Source_Organism,                    ~DIAMOND_PctID, ~DIAMOND_Evalue,  ~Foldseek_Description,
  "soda_gh1",      "Roseinatronobacter (Pseudomonadota)",  97.5,       "0.0",             "2J75 GH1 \u03b2-glucosidase (T. maritima)",

  "rumen_gh1",     "Scatonaster (Bacillota)",              84.3,       "1.5e-312",        "4BCE GH1 \u03b2-glycosidase (P. polymyxa)",
  "soda_gh3",      "Limnospira (Cyanobacteriota)",        100.0,       "0.0",             "3SQL GH31 glycosidase (Synechococcus sp.)",
  "rumen_gh3",     "Fibrobacter (Fibrobacterota)",         98.2,       "0.0",             "5XXN GH3 \u03b2-glucosidase (Fibrobacter sp.)",
  "soda_gh5_11",   "CSBr16-58 (Bacteroidota)",            50.9,       "9.6e-155",        "1WKA Val-tRNA synthetase (T. thermophilus)*",
  "rumen_gh5_11",  "UBA1067 (Verrucomicrobiota)",          52.9,       "6.6e-116",        "1GAX Val-tRNA synthetase (T. thermophilus)*",
  "soda_gh9",      "Limnospira (Cyanobacteriota)",        100.0,       "0.0",             "3X17 GH9 endoglucanase (metagenome)",
  "rumen_gh9",     "Fibrobacter (Fibrobacterota)",         99.5,       "0.0",             "6DHT GH9 endoglucanase (Ruminiclostridium)",
  "soda_gh10",     "Aquiflexum (Bacteroidota)",            71.2,       "3.2e-188",        "4K68 GH10 xylanase (metagenome)",
  "rumen_gh10",    "Fibrobacter (Fibrobacterota)",         99.1,       "0.0",             "7D88 GH10 xylanase (Bacillus sp. KW1)",
  "soda_gh28",     "Pararhodobacter (Pseudomonadota)",     67.2,       "5.1e-120",        "5EJK RSV integrase (H. sapiens)*",
  "rumen_gh28",    "Prevotella (Bacteroidota)",            94.7,       "0.0",             "3JUR GH28 exoPGase (Bacteroides sp.)"
)

# ── Merge all data sources ────────────────────────────────────────────────────
table1_df <- struct_df %>%
  select(Structure, Environment, GH_Family, LDDT, pTM) %>%
  left_join(af3_summary %>% select(Structure, Mean_pLDDT), by = "Structure") %>%
  left_join(active_summary, by = "Structure") %>%
  left_join(curated_meta, by = "Structure") %>%
  mutate(
    GH_Family   = factor(GH_Family, levels = c("GH1","GH3","GH5 sub11","GH9","GH10","GH28")),
    Environment = factor(Environment, levels = env_levels)
  ) %>%
  arrange(GH_Family, Environment)

# ── Print summary statistics ──────────────────────────────────────────────────
cat(sprintf("**Structural analysis summary:**\n\n"))
cat(sprintf("- Total structures predicted: %d (6 GH families \u00d7 2 environments)\n", nrow(table1_df)))
cat(sprintf("- pTM range: %.2f\u2013%.2f\n", min(table1_df$pTM), max(table1_df$pTM)))
cat(sprintf("- Foldseek LDDT > 0.7: %d of %d\n",
            sum(table1_df$LDDT > 0.7), nrow(table1_df)))
cat(sprintf("- Active-site residues conserved: %d of %d\n",
            sum(active_sites$Conserved == "True" | active_sites$Conserved == TRUE),
            nrow(active_sites)))

n_struct_with_conserved <- active_sites %>%
  filter(Conserved == "True" | Conserved == TRUE) %>%
  pull(Structure) %>% n_distinct()
cat(sprintf("- Structures with \u2265 1 conserved catalytic residue: %d of %d\n",
            n_struct_with_conserved, n_distinct(active_sites$Structure)))

conserved_plddt <- active_sites %>%
  filter((Conserved == "True" | Conserved == TRUE) & pLDDT_at_site != "N/A") %>%
  mutate(plddt_num = as.numeric(pLDDT_at_site))
cat(sprintf("- pLDDT at conserved active sites: %.1f\u2013%.1f\n",
            min(conserved_plddt$plddt_num), max(conserved_plddt$plddt_num)))

gh28_soda <- table1_df %>% filter(Structure == "soda_gh28")
cat(sprintf("- GH28 soda lake outlier: pTM=%.2f, LDDT=%.3f, DIAMOND %%ID=%.1f%%\n",
            gh28_soda$pTM, gh28_soda$LDDT, gh28_soda$DIAMOND_PctID))

# Non-CAZyme Foldseek hits
cat(sprintf("- Non-CAZyme Foldseek PDB hits (marked *): GH5_11 soda, GH5_11 rumen, GH28 soda\n"))
cat(sprintf("  These share ancestral fold scaffolds despite non-CAZyme top PDB matches\n"))

# ── Format display table ─────────────────────────────────────────────────────
table1_display <- table1_df %>%
  transmute(
    `GH Family`          = GH_Family,
    `Env.`               = Environment,
    `Source Organism (Phylum)` = Source_Organism,
    `DIAMOND %ID`        = sprintf("%.1f", DIAMOND_PctID),
    `DIAMOND E-value`    = DIAMOND_Evalue,
    `pTM`                = sprintf("%.2f", pTM),
    `Mean pLDDT`         = sprintf("%.1f", Mean_pLDDT),
    `Catalytic Residues` = residue_str,
    `Active Site pLDDT`  = plddt_str,
    `Top Foldseek PDB Hit` = Foldseek_Description,
    `LDDT`               = sprintf("%.2f", LDDT)
  )

kable(table1_display, format = "html", escape = FALSE,
      caption = paste0(
        "Table 1. AlphaFold3 structural predictions, dbCAN/DIAMOND CAZyme annotations, ",
        "active-site conservation, and Foldseek PDB100 structural similarity for twelve ",
        "representative CAZymes from soda lake and rumen metagenomes. ",
        "dbCAN Tools: all proteins annotated as CAZymes by DIAMOND against the CAZy database. ",
        "The Foldseek column shows the closest whole-structure match in PDB100; 9 of 12 proteins ",
        "matched genuine CAZymes; the 3 non-CAZyme matches (GH5_11, soda GH28, marked with *) ",
        "share ancestral fold scaffolds. Source organisms of PDB hits shown in parentheses."
      )) %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "hover"), font_size = 8,
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#E8E8E8") %>%
  row_spec(which(table1_df$Structure == "soda_gh28"), background = "#FFF3CD") %>%
  column_spec(3, width = "12em") %>%
  column_spec(8, width = "10em") %>%
  column_spec(10, width = "14em") %>%
  footnote(
    general = c(
      "* Non-CAZyme Foldseek hit; shares ancestral fold scaffold with target GH family.",
      "Catalytic residues shown as predicted amino acid + position; Variant = substitution at expected catalytic site.",
      "DIAMOND %ID and E-value from CAZy database annotation via dbCAN run_dbcan pipeline."
    ),
    general_title = "Notes: "
  )
```

## 7.8 Table 2 — Extremophile-Derived CAZymes as Industrial Biocatalysts

```{r table2-biotech, results='asis'}
# ── Build Table 2: Biotechnological potential of predicted CAZymes ─────────
# Combines structural confidence, active-site conservation, and literature-
# informed industrial applications for each GH family × environment pair.

# Industrial application annotations (literature-based)
biotech_annotations <- tibble::tribble(
  ~GH_Family,   ~Substrate,                    ~Reaction,                          ~Industrial_Application,                                           ~Optimal_Conditions,
  "GH1",        "Cellobiose, lactose, glucosides", "Hydrolysis of terminal β-glucosidic bonds", "Cellulosic ethanol (SSF), food flavour release, pharmaceutical glycoside activation", "Alkaliphilic variants: pH 9–11; Thermophilic: 60–80 °C",
  "GH3",        "Cellobiose, β-glucans, xylan",    "Exo-β-glucosidase / β-xylosidase",         "Biomass saccharification, animal feed enzyme cocktails, biofuel production",           "Broad pH range; mesophilic to moderately thermophilic",
  "GH5 sub11",  "β-1,4-glucan (CMC, barley glucan)","Endo-β-1,4-glucanase",                   "Textile bio-polishing, detergent formulations, pulp & paper processing",               "Alkaliphilic: pH 8–11 (soda lake); Neutral: pH 6–7 (rumen)",
  "GH9",        "Crystalline & amorphous cellulose","Processive endoglucanase",                 "Consolidated bioprocessing, biomass pretreatment, textile finishing",                   "Thermostable variants preferred for industrial CBP",
  "GH10",       "Xylan, arabinoxylan",              "Endo-β-1,4-xylanase",                     "Paper bleaching, bread-making, animal feed, xylooligosaccharide production",           "Alkaliphilic: detergent/pulp; Acidic: food/feed applications",
  "GH28",       "Pectin, polygalacturonan",         "Polygalacturonase",                        "Fruit juice clarification, textile retting, wine processing, bio-scouring",            "Acidic pH for juice; alkaline pH for textile applications"
)

# Merge structural data with biotech annotations
table2_df <- struct_df %>%
  select(Structure, Environment, GH_Family, pTM, Seq_Identity, LDDT) %>%
  left_join(
    af3_summary %>% select(Structure, Mean_pLDDT),
    by = "Structure"
  ) %>%
  left_join(
    active_sites %>%
      group_by(Structure) %>%
      summarise(
        n_sites     = n(),
        n_conserved = sum(Conserved == "True" | Conserved == TRUE),
        .groups     = "drop"
      ),
    by = "Structure"
  ) %>%
  left_join(biotech_annotations, by = "GH_Family") %>%
  mutate(
    GH_Family   = factor(GH_Family, levels = c("GH1","GH3","GH5 sub11","GH9","GH10","GH28")),
    Environment = factor(Environment, levels = env_levels),
    # Structural readiness score: high pTM + high pLDDT + conserved active sites
    Structural_Readiness = case_when(
      pTM >= 0.90 & Mean_pLDDT >= 85 & n_conserved == n_sites ~ "High",
      pTM >= 0.80 & Mean_pLDDT >= 70 & n_conserved >= 1       ~ "Moderate",
      TRUE                                                      ~ "Low"
    ),
    Conserved_str = paste0(n_conserved, "/", n_sites)
  ) %>%
  arrange(GH_Family, Environment)

# Display table
table2_display <- table2_df %>%
  select(GH_Family, Environment, Substrate, Reaction, Industrial_Application,
         pTM, Mean_pLDDT, Conserved_str, Structural_Readiness) %>%
  mutate(
    pTM       = sprintf("%.2f", pTM),
    Mean_pLDDT = sprintf("%.1f", Mean_pLDDT)
  ) %>%
  rename(
    `GH Family`       = GH_Family,
    `Env.`            = Environment,
    `Substrate`       = Substrate,
    `Reaction`        = Reaction,
    `Industrial Application` = Industrial_Application,
    `pTM`             = pTM,
    `Mean pLDDT`      = Mean_pLDDT,
    `Active Sites`    = Conserved_str,
    `Readiness`       = Structural_Readiness
  )

# Colour-code readiness
readiness_colours <- ifelse(table2_df$Structural_Readiness == "High", "#C3E6CB",
                     ifelse(table2_df$Structural_Readiness == "Moderate", "#FFEEBA", "#F5C6CB"))

kable(table2_display, format = "html", escape = FALSE,
      caption = "Table 2. Biotechnological potential of extremophile-derived CAZymes: structural confidence and industrial applications for twelve AlphaFold3-predicted enzymes from Ethiopian soda lake and rumen metagenomes.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), font_size = 8,
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#E8E8E8") %>%
  column_spec(9, background = readiness_colours, bold = TRUE) %>%
  footnote(
    general = c(
      "Structural Readiness: High = pTM ≥ 0.90, pLDDT ≥ 85, all active sites conserved; Moderate = pTM ≥ 0.80, pLDDT ≥ 70, ≥ 1 conserved; Low = below thresholds.",
      "Industrial applications compiled from CAZy database and reviewed literature (Lombard et al. 2014; Jeilu et al. 2024)."
    ),
    general_title = "Notes: "
  )

# Summary stats
cat("\n\n**Biotechnology readiness summary:**\n\n")
cat(sprintf("- High readiness candidates: %d of %d\n",
            sum(table2_df$Structural_Readiness == "High"), nrow(table2_df)))
cat(sprintf("- Moderate readiness: %d of %d\n",
            sum(table2_df$Structural_Readiness == "Moderate"), nrow(table2_df)))
cat(sprintf("- Low readiness (require further validation): %d of %d\n",
            sum(table2_df$Structural_Readiness == "Low"), nrow(table2_df)))
cat(sprintf("- Soda lake enzymes offer alkali/salt-tolerant variants for detergent & textile industries\n"))
cat(sprintf("- Rumen enzymes offer efficient biomass-degrading variants for biofuel & feed applications\n"))
```

## 7.9 ★ Figure 7: AlphaFold3-Predicted CAZyme Structures (6×2 Grid)


```{r fig6-composite, fig.height=21, fig.width=16}
# ── Figure 7: Individual predicted structures coloured by pLDDT ──────────────
# Columns: Soda Lake (left) | Rumen (right)
# Rows:    GH1, GH3, GH5_11, GH9, GH10, GH28
# Wider canvas (16" × 21") so each panel has enough horizontal room for
# ribbons + catalytic-residue labels to render clearly.
p_renders
save_fig(p_renders, "Fig7_structural_analysis", 16, 21)
```

## 7.10 Supplementary Figure S3 — Structural Superposition and Conservation

```{r figS3-extended-structural, fig.height=18, fig.width=12.60}
# ── Supplementary Figure S3 ──────────────────────────────────────────────────
# A: Structural superpositions + active-site close-ups (from PyMOL)
# B: Per-residue pLDDT confidence profiles
# C: Sequence identity vs TM-score scatter
# D: Active-site conservation tiles

p_sup_grob <- wrap_elements(full = grid::grid.grabExpr(print(p_superpositions)))

p_figS3 <- p_sup_grob /
            (p_plddt + labs(title = NULL)) /
            (p_struct_scatter + labs(title = NULL, subtitle = NULL) |
             p_active_site + labs(title = NULL)) +
  plot_layout(heights = c(2, 1, 1)) +
  plot_annotation(
    tag_levels = "A",
    title = "Structural superposition, pLDDT profiles, and active-site conservation"
  ) &
  theme_pub() &
  theme(plot.tag = element_text(size = 10, face = "bold"),
        plot.margin = margin(4, 4, 4, 4))
p_figS3
save_fig(p_figS3, "FigS3_extended_structural", nm_wide, 18)
```

---

# 9 — Manuscript Statistics Summary 
This section prints all key statistics needed for the Results section of the manuscript.

## 9.1 Sample & MAG Counts

```{r ms-counts}
cat("═══════════════════════════════════════════════════════════\n")
cat("SAMPLE & MAG COUNTS\n")
cat("═══════════════════════════════════════════════════════════\n\n")

n_samples <- ncol(sp_mat)
n_sl_samples <- sum(mpa_samples$Environment == "Soda Lake")
n_ru_samples <- sum(mpa_samples$Environment == "Rumen")
cat(sprintf("Total metagenomic samples: %d (%d soda lake, %d rumen)\n",
            n_samples, n_sl_samples, n_ru_samples))

cat("\nSamples per source:\n")
print(table(mpa_samples$Source, mpa_samples$Environment))

n_mags_total <- nrow(red_df)
n_mags_sl <- sum(red_df$Environment == "Soda Lake")
n_mags_ru <- sum(red_df$Environment == "Rumen")
cat(sprintf("\nQuality-filtered MAGs (with RED): %d (%d soda lake, %d rumen)\n",
            n_mags_total, n_mags_sl, n_mags_ru))

# Also count from novelty_rank (full GTDB-Tk set)
if (exists("novelty_rank")) {
  cat(sprintf("Total GTDB-Tk classified MAGs: %d (%d SL, %d RU)\n",
              nrow(novelty_rank),
              sum(novelty_rank$Environment == "Soda Lake"),
              sum(novelty_rank$Environment == "Rumen")))
}

# Prokka-annotated MAGs
n_prokka_mags <- n_distinct(prokka_ann$MAG)
n_prokka_sl <- n_distinct(prokka_ann$MAG[prokka_ann$Environment == "Soda Lake"])
n_prokka_ru <- n_distinct(prokka_ann$MAG[prokka_ann$Environment == "Rumen"])
cat(sprintf("Prokka-annotated MAGs: %d (%d SL, %d RU)\n",
            n_prokka_mags, n_prokka_sl, n_prokka_ru))

cat(sprintf("Total Prokka gene annotations: %s\n", format(nrow(prokka_ann), big.mark = ",")))
```

## 9.2 MetaPhlAn — Top Taxa (Phylum & Genus)

```{r ms-taxa-metaphlan}
cat("═══════════════════════════════════════════════════════════\n")
cat("METAPHLAN — TOP TAXA\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# Top phyla per environment
cat("── Top 10 Phyla by mean abundance per environment ──\n\n")
phy_by_env <- mpa_phy_long %>%
  group_by(Phylum, Environment) %>%
  summarise(mean_abd = mean(Abundance), .groups = "drop")

cat("SODA LAKE:\n")
phy_by_env %>% filter(Environment == "Soda Lake") %>%
  arrange(desc(mean_abd)) %>% head(10) %>%
  mutate(label = sprintf("  %2d. %-25s  mean = %.2f%%", row_number(), Phylum, mean_abd)) %>%
  pull(label) %>% cat(sep = "\n")

cat("\n\nRUMEN:\n")
phy_by_env %>% filter(Environment == "Rumen") %>%
  arrange(desc(mean_abd)) %>% head(10) %>%
  mutate(label = sprintf("  %2d. %-25s  mean = %.2f%%", row_number(), Phylum, mean_abd)) %>%
  pull(label) %>% cat(sep = "\n")

# Top genera per environment
cat("\n\n── Top 15 Genera by mean abundance per environment ──\n\n")
gen_by_env <- mpa_genus_clean %>%
  select(-clade_name, -depth) %>%
  pivot_longer(-Genus, names_to = "Sample", values_to = "Abundance") %>%
  left_join(mpa_samples, by = "Sample") %>%
  group_by(Genus, Environment) %>%
  summarise(mean_abd = mean(Abundance), .groups = "drop")

cat("SODA LAKE:\n")
gen_by_env %>% filter(Environment == "Soda Lake") %>%
  arrange(desc(mean_abd)) %>% head(15) %>%
  mutate(label = sprintf("  %2d. %-30s  mean = %.2f%%", row_number(), Genus, mean_abd)) %>%
  pull(label) %>% cat(sep = "\n")

cat("\n\nRUMEN:\n")
gen_by_env %>% filter(Environment == "Rumen") %>%
  arrange(desc(mean_abd)) %>% head(15) %>%
  mutate(label = sprintf("  %2d. %-30s  mean = %.2f%%", row_number(), Genus, mean_abd)) %>%
  pull(label) %>% cat(sep = "\n")
```

## 9.3 Diversity & Statistical Tests

```{r ms-diversity-stats}
cat("═══════════════════════════════════════════════════════════\n")
cat("DIVERSITY & STATISTICAL TESTS\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# MetaPhlAn alpha diversity
cat("── MetaPhlAn Alpha Diversity (species-level) ──\n")
sl_sha <- mpa_alpha$Shannon[mpa_alpha$Environment == "Soda Lake"]
ru_sha <- mpa_alpha$Shannon[mpa_alpha$Environment == "Rumen"]
cat(sprintf("  Shannon — Soda Lake: median = %.2f (IQR: %.2f–%.2f)\n",
            median(sl_sha), quantile(sl_sha, 0.25), quantile(sl_sha, 0.75)))
cat(sprintf("  Shannon — Rumen:     median = %.2f (IQR: %.2f–%.2f)\n",
            median(ru_sha), quantile(ru_sha, 0.25), quantile(ru_sha, 0.75)))
cat(sprintf("  Kruskal-Wallis (Shannon ~ Env): chi-sq = %.2f, p = %.2e\n",
            kw_sha$statistic, kw_sha$p.value))
cat(sprintf("  Kruskal-Wallis (Observed ~ Env): chi-sq = %.2f, p = %.2e\n",
            kw_obs$statistic, kw_obs$p.value))

# MetaPhlAn PERMANOVA
cat(sprintf("\n  PERMANOVA (Bray-Curtis, MetaPhlAn): R² = %.3f, p = %.3f\n",
            perm_mpa$R2[1], perm_mpa$`Pr(>F)`[1]))

# CAZyme alpha diversity
cat("\n── CAZyme Alpha Diversity ──\n")
sl_caz_sha <- alpha_div$Shannon[alpha_div$Environment == "Soda Lake"]
ru_caz_sha <- alpha_div$Shannon[alpha_div$Environment == "Rumen"]
cat(sprintf("  Shannon — Soda Lake: median = %.2f\n", median(sl_caz_sha)))
cat(sprintf("  Shannon — Rumen:     median = %.2f\n", median(ru_caz_sha)))
cat(sprintf("  Kruskal-Wallis (CAZyme Shannon ~ Env): p = %.2e\n", kw_shan$p.value))
cat(sprintf("  Kruskal-Wallis (CAZyme Richness ~ Env): p = %.2e\n", kw_rich$p.value))

# CAZyme PERMANOVA
cat(sprintf("\n  PERMANOVA (Bray-Curtis, CAZyme families): R² = %.3f, p = %.3f\n",
            perm_caz$R2[1], perm_caz$`Pr(>F)`[1]))
cat(sprintf("  NMDS stress: %.3f\n", nmds_caz$stress))

# CAZyme identity
cat("\n── CAZyme Sequence Identity ──\n")
cat(sprintf("  Wilcoxon test (Identity ~ Env): p = %.2e\n", wil_id$p.value))
cat(sprintf("  Median identity — Soda Lake: %.1f%%\n",
            novel_frac$median_id[novel_frac$Environment == "Soda Lake"]))
cat(sprintf("  Median identity — Rumen:     %.1f%%\n",
            novel_frac$median_id[novel_frac$Environment == "Rumen"]))

# RED
cat("\n── RED (Relative Evolutionary Divergence) ──\n")
sl_red <- red_df$RED[red_df$Environment == "Soda Lake"]
ru_red <- red_df$RED[red_df$Environment == "Rumen"]
cat(sprintf("  Soda Lake RED: median = %.3f (n = %d)\n", median(sl_red, na.rm=TRUE), length(sl_red)))
cat(sprintf("  Rumen RED:     median = %.3f (n = %d)\n", median(ru_red, na.rm=TRUE), length(ru_red)))
cat(sprintf("  Kruskal-Wallis (RED ~ Env): p = %.2e\n", kw_env$p.value))

# Unclassified reads
cat(sprintf("\n── Unclassified Reads ──\n"))
cat(sprintf("  Wilcoxon test (%%Unclassified ~ Env): p = %.3e\n", wil_uc$p.value))
```

## 9.4 MAG-Level Taxonomy (GTDB-Tk)

```{r ms-taxa-mags}
cat("═══════════════════════════════════════════════════════════\n")
cat("MAG-LEVEL TAXONOMY (GTDB-Tk)\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# Use novelty_rank which already has parsed taxonomy (d, p, c, o, f, g, s)
cat(sprintf("Total MAGs with GTDB-Tk classification: %d\n\n", nrow(novelty_rank)))

# Top phyla per environment
cat("── Top Phyla in MAGs ──\n\n")
mag_phy <- novelty_rank %>%
  filter(!is.na(p), p != "") %>%
  count(p, Environment) %>%
  group_by(Environment) %>% arrange(desc(n)) %>% slice_head(n = 10)

cat("SODA LAKE:\n")
mag_phy %>% filter(Environment == "Soda Lake") %>% ungroup() %>%
  mutate(label = sprintf("  %2d. %-30s  n = %d MAGs", row_number(), p, n)) %>%
  pull(label) %>% cat(sep = "\n")

cat("\n\nRUMEN:\n")
mag_phy %>% filter(Environment == "Rumen") %>% ungroup() %>%
  mutate(label = sprintf("  %2d. %-30s  n = %d MAGs", row_number(), p, n)) %>%
  pull(label) %>% cat(sep = "\n")

# Top genera per environment
cat("\n\n── Top Genera in MAGs (classified only) ──\n\n")
mag_gen <- novelty_rank %>%
  filter(!is.na(g), g != "") %>%
  count(g, Environment) %>%
  group_by(Environment) %>% arrange(desc(n)) %>% slice_head(n = 15)

cat("SODA LAKE:\n")
mag_gen %>% filter(Environment == "Soda Lake") %>% ungroup() %>%
  mutate(label = sprintf("  %2d. %-30s  n = %d MAGs", row_number(), g, n)) %>%
  pull(label) %>% cat(sep = "\n")

cat("\n\nRUMEN:\n")
mag_gen %>% filter(Environment == "Rumen") %>% ungroup() %>%
  mutate(label = sprintf("  %2d. %-30s  n = %d MAGs", row_number(), g, n)) %>%
  pull(label) %>% cat(sep = "\n")

# Unclassified genus counts
cat("\n\n── Unclassified at Genus Level ──\n")
unclass_g <- novelty_rank %>% filter(is.na(g) | g == "") %>% count(Environment)
cat(sprintf("  Soda Lake: %d MAGs unclassified at genus\n",
            unclass_g$n[unclass_g$Environment == "Soda Lake"]))
cat(sprintf("  Rumen:     %d MAGs unclassified at genus\n",
            unclass_g$n[unclass_g$Environment == "Rumen"]))

# Species (if any classified)
cat("\n── Species-Level Classifications ──\n")
mag_sp <- novelty_rank %>%
  filter(!is.na(s), s != "") %>%
  count(s, Environment) %>% arrange(desc(n))
cat(sprintf("  Total MAGs with species assignment: %d / %d (%.1f%%)\n",
            sum(mag_sp$n), nrow(novelty_rank), sum(mag_sp$n)/nrow(novelty_rank)*100))
if (nrow(mag_sp) > 0) {
  head(mag_sp, 20) %>%
    mutate(label = sprintf("  %-40s  %s (n = %d)", s, Environment, n)) %>%
    pull(label) %>% cat(sep = "\n")
}

# Novelty summary
cat("\n\n── Novelty Summary ──\n")
cat(sprintf("  Novel species (no species match): %d (%.1f%%)\n",
            sum(novelty_rank$novel_species), mean(novelty_rank$novel_species)*100))
cat(sprintf("  Novel genus (no genus match):     %d (%.1f%%)\n",
            sum(novelty_rank$novel_genus), mean(novelty_rank$novel_genus)*100))
cat(sprintf("  Novel family (no family match):   %d (%.1f%%)\n",
            sum(novelty_rank$novel_family), mean(novelty_rank$novel_family)*100))
```

## 9.5 Top GH Families & Genera

```{r ms-gh-stats}
cat("═══════════════════════════════════════════════════════════\n")
cat("TOP GH FAMILIES & CONTRIBUTING GENERA\n")
cat("═══════════════════════════════════════════════════════════\n\n")

# Total CAZyme count
cat(sprintf("Total CAZyme annotations (2-tool consensus): %s\n",
            format(nrow(cazyme_2t), big.mark = ",")))

# GH family counts
gh_all <- cazyme_2t %>% filter(CAZyme_Class == "GH")
cat(sprintf("Total GH annotations: %s\n\n", format(nrow(gh_all), big.mark = ",")))

cat("── Top 20 GH Families (overall) ──\n")
gh_all %>% count(GH_Family = CAZyme_Family, sort = TRUE) %>% head(20) %>%
  mutate(label = sprintf("  %2d. %-10s  n = %5d", row_number(), GH_Family, n)) %>%
  pull(label) %>% cat(sep = "\n")

cat("\n\n── Top 10 GH Families per Environment ──\n\n")
cat("SODA LAKE:\n")
gh_all %>% filter(Environment == "Soda Lake") %>%
  count(Family = CAZyme_Family, sort = TRUE) %>% head(10) %>%
  mutate(label = sprintf("  %2d. %-10s  n = %d", row_number(), Family, n)) %>%
  pull(label) %>% cat(sep = "\n")

cat("\n\nRUMEN:\n")
gh_all %>% filter(Environment == "Rumen") %>%
  count(Family = CAZyme_Family, sort = TRUE) %>% head(10) %>%
  mutate(label = sprintf("  %2d. %-10s  n = %d", row_number(), Family, n)) %>%
  pull(label) %>% cat(sep = "\n")

# GH family enrichment (Fisher's exact)
cat("\n\n── GH Family Enrichment (Fisher's exact, BH-adjusted) ──\n")
cat("  (Showing families with p_adj < 0.05)\n\n")
if (exists("gh_fisher")) {
  sig_gh <- gh_fisher %>% filter(p_adj < 0.05) %>% arrange(p_adj)
  if (nrow(sig_gh) > 0) {
    sig_gh %>%
      mutate(direction = ifelse(log2_OR > 0, "SL-enriched", "RU-enriched"),
             label = sprintf("  %-10s  log2OR = %+.2f  p_adj = %.2e  %s",
                             CAZyme_Family, log2_OR, p_adj, direction)) %>%
      pull(label) %>% cat(sep = "\n")
  } else { cat("  No significantly enriched GH families at p_adj < 0.05\n") }
}

# Top genera per GH family (uses gh_genus which has Genus from GTDB-Tk join)
cat("\n\n── Top 5 Genera per Major GH Family per Environment ──\n")
if (exists("gh_genus")) {
  for (fam in c("GH1", "GH2", "GH3", "GH5", "GH9", "GH13", "GH28", "GH43")) {
    cat(sprintf("\n  %s:\n", fam))
    gh_gen <- gh_genus %>%
      filter(CAZyme_Family == fam, !is.na(Genus), Genus != "") %>%
      count(Genus, Environment, sort = TRUE) %>%
      group_by(Environment) %>% slice_head(n = 5)
    for (env in c("Soda Lake", "Rumen")) {
      sub <- gh_gen %>% filter(Environment == env)
      if (nrow(sub) > 0) {
        cat(sprintf("    %s: %s\n", env,
                    paste(sprintf("%s (%d)", sub$Genus, sub$n), collapse = ", ")))
      }
    }
  }
}
```

## 9.6 Functional Pathway Enrichment

```{r ms-pathway-enrichment}
cat("═══════════════════════════════════════════════════════════\n")
cat("FUNCTIONAL PATHWAY ENRICHMENT (Figure 4)\n")
cat("═══════════════════════════════════════════════════════════\n\n")

if (exists("enrich_pathway")) {
  enrich_pathway %>%
    arrange(pathway_group, p_adj) %>%
    mutate(
      direction = ifelse(log2_OR > 0, "SL-enriched", "RU-enriched"),
      label = sprintf("  %-35s  %-18s  log2OR = %+6.2f  p_adj = %.2e  %s  %s",
                      as.character(pathway), as.character(pathway_group),
                      log2_OR, p_adj, sig_label, direction)
    ) %>%
    pull(label) %>% cat(sep = "\n")
}

```

## 9.7 AlphaFold3 & Structural Analysis

```{r ms-alphafold-stats}
cat("═══════════════════════════════════════════════════════════\n")
cat("ALPHAFOLD3 STRUCTURE PREDICTIONS\n")
cat("═══════════════════════════════════════════════════════════\n\n")

af3 <- read_tsv(file.path(data_dir, "alphafold3_summary.tsv"), show_col_types = FALSE)
cat("── Per-Structure Summary ──\n\n")
af3 %>%
  mutate(label = sprintf("  %-45s  Env=%-10s  GH=%-5s  Len=%4d  pTM=%.2f  meanPLDDT=%.1f  HiConf=%.1f%%",
                          Structure, Environment, GH_Family, Sequence_Length,
                          pTM, Mean_pLDDT, High_Confidence_Pct)) %>%
  pull(label) %>% cat(sep = "\n")

cat("\n\n── Foldseek Best Hits (Sequence Identity & LDDT) ──\n\n")
fs_hits <- read_tsv(file.path(data_dir, "foldseek_best_hits.tsv"), show_col_types = FALSE)
fs_hits %>%
  mutate(label = sprintf("  %-20s  PDB=%-30s  SeqID=%.1f%%  LDDT=%.3f",
                          Query, Best_PDB_Hit, Seq_Identity*100, LDDT)) %>%
  pull(label) %>% cat(sep = "\n")

cat("\n\n── Structural Superposition RMSD (SL vs RU pairs, TM-align) ──\n")
cat("  GH1:    RMSD = 2.71 A  (TM-score = 0.90, 425 aligned)\n")
cat("  GH3:    RMSD = 3.61 A  (TM-score = 0.76, 457 aligned)\n")
cat("  GH5_11: RMSD = 1.78 A  (TM-score = 0.91, 337 aligned)\n")
cat("  GH9:    RMSD = 2.52 A  (TM-score = 0.80, 527 aligned)\n")
cat("  GH10:   RMSD = 2.87 A  (TM-score = 0.67, 302 aligned)\n")
cat("  GH28:   RMSD = 7.11 A  (TM-score = 0.25, 147 aligned)\n")

cat("\n\n── Active Site Conservation ──\n\n")
active <- read_tsv(file.path(data_dir, "renders/active_site_conservation.tsv"),
                   show_col_types = FALSE) %>%
  mutate(
    Pred_Position  = as.integer(Pred_Position),
    pLDDT_at_site  = as.numeric(na_if(pLDDT_at_site, "N/A")),
    Conserved_bool = (Conserved == "True")
  )
active %>%
  mutate(label = sprintf("  %-5s  %-10s  %-45s  Role=%-15s  Ref=%s  Pred=%s%s  Conserved=%s  pLDDT=%s",
                          GH_Family, Environment, Structure, Role,
                          Ref_Residue, Pred_Residue,
                          ifelse(is.na(Pred_Position), "", as.character(Pred_Position)),
                          Conserved,
                          ifelse(is.na(pLDDT_at_site), "N/A", sprintf("%.1f", pLDDT_at_site)))) %>%
  pull(label) %>% cat(sep = "\n")

n_conserved <- sum(active$Conserved_bool, na.rm = TRUE)
n_total_sites <- nrow(active)
cat(sprintf("\n\n  TOTAL: %d/%d catalytic residues conserved (%.0f%%)\n",
            n_conserved, n_total_sites, n_conserved/n_total_sites*100))
cat(sprintf("  Min pLDDT at active site: %.1f\n", min(active$pLDDT_at_site, na.rm = TRUE)))
cat(sprintf("  Max pLDDT at active site: %.1f\n", max(active$pLDDT_at_site, na.rm = TRUE)))
```

---

# Session Info

```{r session-info}
sessionInfo()
```
